---
title: "**secr** simulations Mt"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Temporal variation in detection parameters Mt

The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

Temporal variation is generally not thought to be a problem for SECR estimates of density (Sollmann 2024). These simulations extend the empirical basis for the conclusion.

## Problem description

The SECR null model assumes $\lambda_0$ and $\sigma$ are constant over occasions. We vary these parameters and examine the resulting estimates from a null model.

## Settings

Square grid of 64 binary proximity detectors at spacing $c = 2\sigma$.  

The same conditions were used for simulation and fitting. 

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| D | $12.5 \sigma^{-2}$ |
| lambda0  | 0.1 |
| sigma    | c/2 |
| noccasions | 10 |
| buffer   | $4\sigma$ |
| nx       | 32 |

## Package version, date and platform


```{r versions, results = 'hold', echo = FALSE}
source('../setup.R')
runsim <- TRUE
metadata(runsim) # if FALSE picks up saved metadata.RDS
```
## Simulation code

### Custom function to simulate capture histories

We vary the time-specific detection parameters at each replicate, using a log-normal random variate to keep values positive.

```{r Mt capthist}
source('../setup.R')
simCV.capthist <- function (detectpar = list(), CVlambda0 = 0, CVsigma =0, ...)
{
    nocc <- list(...)$noccasions
    if (CVlambda0>0) 
        detectpar$lambda0 <- secr::rlnormCV(nocc, detectpar$lambda0, CVlambda0)
    if (CVsigma>0) 
        detectpar$sigma <- secr::rlnormCV(nocc, detectpar$sigma, CVsigma)
    sim.capthist(detectpar = detectpar, ...)
}
```

Visualising the parameter distributions:

```{r Mt demo}
par(mfrow=c(1,2))
cvlevels <- c(0.1, 0.2, 0.3, 0.4)
lam0 <- lapply(cvlevels, secr::rlnormCV, n = 10000, mean = 0.1)
sig <- lapply(cvlevels, secr::rlnormCV, n = 10000, mean = 20)
boxplot(lam0, ylim=c(0,0.4), names = cvlevels, xlab = 'CV', ylab = 'lambda0')
boxplot(sig, ylim=c(0,60), names = cvlevels, xlab = 'CV', ylab = 'sigma')
```

### Define Mt scenarios

```{r Mt scenarios, eval = TRUE, message = FALSE, results = "hide"}
grid <- make.grid(nx = 8, ny = 8, spacing = 40, detector = "proximity")
detarg <- c(expand.arg(CVlambda0 = seq(0,0.4,0.1), CVsigma = 0),
            expand.arg(CVlambda0 = 0, CVsigma = seq(0.1,0.4,0.1)))
fitarg <- expand.arg(detectfn = 'HHN', model = c(lambda0 ~ 1,lambda0 ~ t))
# Mean lambda0 and sigma in scenario (used as reference for RB etc.)
# CV is provided by detarg and used by simCV.capthist
# ignore model ~t for now
scen <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, 
    lambda0 = 0.1, sigma = 20, detindex = 1:length(detarg), fitindex = 1)
```

```{r Mt scenario summary, cache = TRUE}
scenarioSummary(scen, grid)
```

### Check empirical CV of counts per occasion

```{r Mt counts, echo = TRUE, eval = TRUE, cache = TRUE, message = FALSE}
Mtraw <- run.scenarios(nrepl = 100, scen, trapset = grid, det.args = detarg, 
    CH.function = "simCV.capthist", fit = FALSE, seed = 12345, extractfn = identity)
# average CV(detections) for each scenario
cvt <- function(y) CV(apply(y,2,sum))  # temporal CV(detections)
CVmat <- sapply(Mtraw$output, function (x) sapply(x, cvt))
apply(CVmat,2,mean)
```

### Run Mt simulations

```{r Mt run, eval = runsim}
Mtsims <- run.scenarios(nrepl = 200, scen, trapset = grid, det.args = detarg, 
    CH.function = "simCV.capthist", fit = TRUE, fit.args = fitarg, seed = 12345)
saveRDS(Mtsims, file = 'Mtsims.RDS')
```

## Results

```{r Mt tables, eval = TRUE, echo = FALSE, results = "hold"}
options(width=110, digits=4)
Mtsims <- readRDS(file = 'Mtsims.RDS')
estD <- estimateSummary(Mtsims)[,-1]
estL <- estimateSummary(Mtsims, 'lambda0')[,-1]
estS <- estimateSummary(Mtsims, 'sigma')[,-1]
cvl <- c(0,0.1,0.2,0.3,0.4,0,0,0,0)
cvs <- c(0,0,0,0,0,0.1,0.2,0.3,0.4)
cat('Density\n')
cbind(CVlambda0 = cvl, CVsigma = cvs, estD)
cat('lambda0\n') 
cbind(CVlambda0 = cvl, CVsigma = cvs, estL)
cat('sigma\n')
cbind(CVlambda0 = cvl, CVsigma = cvs, estS)
```
 
```{r Mt figure, eval = TRUE, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Relative bias of parameter estimates from null model given temporal variation in lambda0 or sigma" }
Mtsims <- readRDS(file = 'Mtsims.RDS')
estD <- estimateSummary(Mtsims)[,-1]
estL <- estimateSummary(Mtsims, 'lambda0')[,-1]
estS <- estimateSummary(Mtsims, 'sigma')[,-1]
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o')
x <- seq(0,0.4,0.1)

plot(0,0,type='n', xlim=c(0,0.4), ylim=c(-0.4,0.4), xlab = 'CV(lambda0)', ylab='RB')
abline(h=0, lty=2)
addRB(x, estD[1:5,], type='o', pch=16)
addRB(x, estL[1:5,], type='o', pch=21, bg='white')
addRB(x, estS[1:5,], type='o', pch=24, bg='white')
legend(0.01, 0.39, legend=c('D','lambda0','sigma'), pch=c(16,1,2))

plot(0,0,type='n', xlim=c(0,0.4), ylim=c(-0.4,0.4), xlab = 'CV(sigma)', ylab='RB')
abline(h=0, lty=2)
addRB(x, estD[c(1,6:9),], type='o', pch=16)
addRB(x, estL[c(1,6:9),], type='o', pch=21, bg='white')
addRB(x, estS[c(1,6:9),], type='o', pch=24, bg='white')
legend(0.01, 0.39, legend=c('D','lambda0','sigma'), pch=c(16,1,2))
```

## Interpretation

Estimates of density from the null model are indeed robust to temporal variation in $\lambda_0$ or $\sigma$, but variation in $\sigma$ drives complementary bias in $\hat \lambda_0$ and $\hat \sigma$

## One-off

Run this code to obtain results for a single scenario (here CV(lambda0) = 1). Re-uses code in chunk 'Mt capthist'.
```{r MT oneoff, eval = TRUE, cache = TRUE, message = FALSE, warning = FALSE}
<<Mt capthist>> 
grid <- make.grid(nx = 8, ny = 8, spacing = 40, detector = "proximity")
detarg <- list(detectpar = list(CVlambda0 = 1, CVsigma = 0))
fitarg <- list(detectfn = 'HHN')  # null model
scen <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, lambda0 = 0.1, sigma = 20)
oneoff <- run.scenarios(nrepl = 100, scen, trapset = grid, det.args = detarg, 
    CH.function = "simCV.capthist", fit = TRUE, fit.args = fitarg, seed = 12345)
summary(oneoff)
```
