---
title: "**secr** simulations DNC"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Dispersal and non-closure DNC

The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

## Problem description

Non-closure due to immigration and emigration (movement of AC) during the sampling period.

## Settings

Square grid of 64 binary proximity detectors at 3000-m spacing.  

The same conditions used for simulation and fitting. 
Weekly 'sessions' were joined in `run.scenarios` for the final single-session analysis.

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| D        | 0.001 (0.1/km^2^) |
| lambda0  | 0.02 per day |
| sigma    | 2000 m |
| lambda   | 1.0 (no trend in density) |
| phi      | 1.0 |
| noccasions | 100 |
| buffer   | 10000 m |

## Package version, date and platform

```{r versions, results = 'hold', echo = FALSE}
source('../setup.R')
runsim <- TRUE
metadata(runsim) # if FALSE picks up saved metadata.RDS
```

## Simulation code

```{r DNC initial, eval = TRUE, message = FALSE, results = "hide"}
source('../setup.R')
grid <- make.grid(nx = 8, ny = 8, spacing = 3000, detector = "proximity")
moves <- c(0, 250, 500, 750, 1000)
nmove <- length(moves)
poparg <- expand.arg(
    lambda    = 1, 
    phi       = 1, 
    move.a    = moves, 
    movemodel = "normal", 
    buffer    = 40000, 
    D         = 0.001, 
    nsessions = 100,
    sublist = list(details = c('lambda','phi','move.a','movemodel')))

detarg <- list(traps = grid, detectfn = 'HHN', detectpar = list(lambda0 = 0.02, sigma = 2000),
               renumber = FALSE, noccasions = 1)
fitarg <- list(mask = make.mask(grid, buffer = 10000, type = 'trapbuffer'), 
               detectfn = 'HHN')
scen <- make.scenarios(D = 0.001, noccasions = 1, detectfn = 'HHN', lambda0 = 0.02, 
                        sigma = 2000, popindex=1:length(moves))
```

```{r DNC code1, eval = runsim}
DNCsims <- run.scenarios(nrepl = 500, scen, trapset = grid, pop.args = popargs, 
                         det.args = detarg, joinsessions = TRUE, fit = TRUE, 
                         fit.args = fitarg, seed = 12345)
saveRDS(DNCsims, file = 'DNCsims.RDS')
```

## Results

```{r DNC results1, echo = FALSE, results = "hold"} 
DNCsims <- readRDS('DNCsims.RDS')
# header(DNCsims)
estD <- estimateSummary(DNCsims)
estL <- estimateSummary(DNCsims, 'lambda0')
estS <- estimateSummary(DNCsims, 'sigma')
counts <- getcounts(DNCsims)
cbind(move.a = moves, counts[,1:4])
countlegend(-(5:6))
```

Runtime for `r DNCsims$nrepl` replicates using 24 cores was `r round(DNCsims$proctime/60)` minutes.

```{r DNCfigure, echo = FALSE, fig.height = 5, fig.width=5, out.width="60%", fig.cap = "Relative bias RB, relative standard error RSE, and coverage of 95% confidence intervals COV for density estimates from studies of varying random movement of AC. Annual survival 50% (open circles), 75% (filled circles) or 100% (triangles). Horizontal lines for reference only."}
leg <- c('D','lambda0','sigma')
par(mfrow = c(1,1), mar = c(5,4,2,2), mgp=c(2.4,0.7,0), pty = 's', xpd = FALSE)
plot(0,0, type = 'n', ylim = c(-1, 1.2), xlim = c(0,1000), 
     xlab = 'Daily HR shift sigma', ylab = 'RB')
abline (h = 0, lty = 2)
addRB(moves, estD, pch = 16, type = 'o')
addRB(moves, estL, pch = 21, bg = 'white', type = 'o')
addRB(moves, estS, pch = 24, bg = 'white', type = 'o')
legend(0,-0.5,leg,pch=c(16,21,24), cex=0.85)
```

## Interpretation

Results are specific to a population with no births or deaths.

## One-off

Run this code to obtain results for a single scenario. Requires secrdesign $\ge$ 2.9.0.
```{r DNC oneoff, eval = FALSE, message = FALSE, warning = FALSE, results = "hide"}
source('../setup.R')
grid   <- make.grid(nx = 8, ny = 8, spacing = 3000, detector = "proximity")
poparg <- list(core = grid, buffer = 40000, D = 0.001, nsessions = 100, 
               details = list(lambda = 1, phi = 1, move.a = 500, movemodel = 'normal'))
detarg <- list(traps = grid, detectfn = 'HHN', detectpar = list(lambda0 = 0.02, sigma = 2000),
               renumber = FALSE, noccasions = 1)
fitarg <- list(mask = make.mask(grid, buffer = 10000, type = 'trapbuffer'), 
               detectfn = 'HHN')
scen   <- make.scenarios( D = 0.001, noccasions = 1, detectfn = 'HHN', lambda0 = 0.02, sigma = 2000)
oneoff <- run.scenarios(nrepl = 1, scen, trapset = grid, pop.args = poparg, det.args = detarg, 
                        joinsessions = TRUE, fit = TRUE, fit.args = fitarg, seed = 12345,
                        extractfn = 'identity')
estimateSummary(oneoff)
```
