---
title: "**secr** simulations Mb"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Behavioural variation in detection parameter $\lambda_0$ Mb

The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

Unmodelled behavioural variation is a potential problem that has not been 
examined to my knowledge. 

## Problem description

The SECR null model assumes $\lambda_0$ and $\sigma$ are not affected by whether an individual has been caught (detected) before. We model a general learned response in $\lambda_0$ and one that is detector-specific, and examine the resulting estimates from a null model.

## Settings

Square grid of 64 binary multi-catch traps at spacing $c = 2\sigma$.  

The same conditions used for simulation and fitting.

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| D | $12.5 \sigma^{-2}$ |
| lambda0  | 0.1 |
| sigma    | c/2 |
| noccasions | 10 |
| buffer   | $4\sigma$ |
| nx       | 32 |

## Package version, date and platform

```{r versions, results = 'hold', echo = FALSE}
source('../setup.R')
runsim <- TRUE
metadata(runsim) # if FALSE picks up saved metadata.RDS
```

## Simulation code

### Define Mb scenarios

```{r Mb scenarios, eval = TRUE, message = FALSE, results = "hide"}
source('../setup.R')
grid <- make.grid(nx = 8, ny = 8, spacing = 40, detector = "multi")
detarg <- expand.arg(
    recapfactor = c(0.25,0.5,1,1.5,2), 
    bk = c(FALSE,TRUE),
    sublist=list(detectpar = c('recapfactor','bk')))
fitarg <- list(detectfn = 'HHN', model = lambda0 ~ 1)
scen <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, 
    lambda0 = 0.1, sigma = 20, detindex = 1:length(detarg), fitindex = 1)
```

```{r Mb scenario summary, cache = TRUE}
scenarioSummary(scen, grid)
```

### Check counts

```{r Mb counts, echo = TRUE, eval = TRUE, cache = TRUE, message = FALSE}
Mbraw <- run.scenarios(nrepl = 100, scen, trapset = grid, det.args = detarg, 
    fit = FALSE, seed = 12345, extractfn = identity)
```

### Run Mb simulations

```{r Mb-run, eval = runsim, warning = FALSE}
Mbsims <- run.scenarios(nrepl = 200, scen, trapset = grid, det.args = detarg, 
    fit = TRUE, fit.args = fitarg, seed = 12345)
saveRDS(Mbsims, file = 'Mbsims.RDS')
```

## Results

```{r Mb tables, eval = TRUE, echo = FALSE, results = "hold"}
options(width=110, digits=4)
Mbsims <- readRDS(file = 'Mbsims.RDS')
scenarios <- t(sapply(detarg, function(x) data.frame(x$detectpar)))
estD <- estimateSummary(Mbsims)[,-1]
estL <- estimateSummary(Mbsims, 'lambda0')[,-1]
estS <- estimateSummary(Mbsims, 'sigma')[,-1]
cat('Density\n')
cbind(scenarios, round(estD,4))
cat('lambda0\n') 
cbind(scenarios, round(estL,4))
cat('sigma\n')
cbind(scenarios, round(estS,4))
```
 
```{r Mb figure, eval = TRUE, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Relative bias of parameter estimates from null model given behavioural variation in lambda0" }
Mbsims <- readRDS(file = 'Mbsims.RDS')
estD <- estimateSummary(Mbsims)[,-1]
estL <- estimateSummary(Mbsims, 'lambda0')[,-1]
estS <- estimateSummary(Mbsims, 'sigma')[,-1]
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o', bg = 'white')
x <- c(0.25,0.5,1,1.5,2)
leg <- c('D','lambda0','sigma')
plot(0,0,type='n', xlim=c(0,2), ylim=c(-0.8,1), xlab = 'Recapture factor', ylab='RB')
abline(h=0, lty=2)
addRB(x, estL[1:5,], type='o', pch=21, bg = 'white')
addRB(x, estS[1:5,], type='o', pch=24, bg = 'white')
addRB(x, estD[1:5,], type='o', pch=16)
legend(1.4, -0.39, legend = leg, pch=c(16,21,24), cex = 0.85)
plot(0,0,type='n', xlim=c(0,2), ylim=c(-0.8,1), xlab = 'Recapture factor', ylab='RB')
abline(h=0, lty=2)
addRB(x, estL[6:10,], type='o', pch=21, bg = 'white')
addRB(x, estS[6:10,], type='o', pch=24, bg = 'white')
addRB(x, estD[6:10,], type='o', pch=16)
legend(1.4, -0.39, legend = leg, pch=c(16,21,24), cex = 0.85)
```

## Interpretation

Estimates of density from the null model are not robust to unmodelled behavioural variation in $\lambda_0$. Bias in the density estimates is much reduced when the learned response is specific to the trap site (bk), but then estimates of $\sigma$ also come to be biased. These results apply to traps and not to proximity detectors for which the behavioural response is poorly defined.

## One-off

Run this code to obtain results for a single scenario (here recapfactor = 2). Here we fit both a null model (lambda0~1) and the generating model (lambda0~b).

```{r Mb oneoff, eval = TRUE, cache = TRUE, message = FALSE, warning = FALSE}
source('../setup.R')
grid <- make.grid(nx = 8, ny = 8, spacing = 40, detector = "multi")
detarg <- list(detectpar = list(recapfactor = 2))
fitarg <- list(
    list(detectfn = 'HHN', model = lambda0~1), 
    list(detectfn = 'HHN', model = lambda0~b) 
)
scen <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, lambda0 = 0.1, 
                       sigma = 20, fitindex = 1:2)
oneoff <- run.scenarios(nrepl = 100, scen, trapset = grid, det.args = detarg, 
    fit = TRUE, fit.args = fitarg, seed = 12345)
round(estimateSummary(oneoff, 'D'),4)
round(estimateSummary(oneoff, 'lambda0'),4)
```