---
title: "**secr** simulations STR"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Stratification and parameter variation STR

The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

Covariation of density and effort or detection leads to bias if not adequately modelled.

<!-- To render from (linux) command line use full file path e.g., -->
<!-- Rscript -e "rmarkdown::render('/home/effmu21p/STR/secr-simulations-STR.rmd')" -->

## Settings

Two square grids of 64 binary proximity detectors at spacing $c = 2$.  

The same conditions were used for simulation and fitting. 

Base scenario

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| D | 12.5 (5,20) |
| lambda0  | 0.1 (0.05,0.15) |
| sigma    | c/2 |
| noccasions | 10 |
| buffer   | $4\sigma$ |
| nx       | 64 |

## Package version, date and platform

```{r versions, results = 'hold', echo = FALSE}
source('../setup.R')
source('stratumTable.R')
runsim <- TRUE
metadata()
```

## Simulation code

### First construct grids

```{r strgrids}
# start with 8 x 8 grids
grid1 <- grid2 <- make.grid(8,8, spacing = 40, detector = 'proximity')
grid2[,1] <- grid2[,1] + 480  # displace to right

covariates(grid1)$grid <- rep(factor(1, levels = 1:2), nrow(grid1))
covariates(grid2)$grid <- rep(factor(2, levels = 1:2), nrow(grid2))

# grids1a is a smaller grid 6 x 6
grid1a <- subset(grid1, !apply(grid1, 1, function(x) any(x<20 | x>260)))

# combine
grids <- rbind(grid1, grid2)
gridsa <- rbind(grid1a, grid2)

# stratify mask horizontally and vertically
mask <- make.mask(grids, buffer = 100, spacing = 10)
covariates(mask)$grid <- factor(1 + (mask$x>380))
covariates(mask)$D <- c(5,20)[covariates(mask)$grid]
covariates(mask)$gridy <- factor(1 + (mask$y>140))
covariates(mask)$Dy <- c(5,20)[covariates(mask)$gridy]

# pops for illustration
popargs <- list(D = 'D', core = mask, model2D = 'IHP')
pop <- do.call(sim.popn, popargs)
popargsy <- list(D = 'Dy', core = mask, model2D = 'IHP')
popy <- do.call(sim.popn, popargsy)
```


### Plot - as in Efford in prep. Study design

```{r strplot, fig.width=8, fig.height=4}
par(mfrow=c(2,3), mar=c(1,1,2,1))

plot(mask, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(pop, pch=16, cex=0.7, col = 'white')
plot(grid1, add=T, detpar=list(pch=16, cex=0.7))
plot(grid2, add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('a. Effort constant, ', hat(italic(D)), ~~'unbiased')), cex=0.9)

plot(mask, cov='gridy', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(popy, pch=16, cex=0.7, col = 'white')
plot(grid1a,add=T, detpar=list(pch=16, cex=0.7))
plot(grid2,add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('b. Effort uncorrelated, ', hat(italic(D)), ~~'unbiased')), cex=0.9)

plot(mask, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(pop, pch=16, cex=0.7, col = 'white')
plot(grid1a,add=T, detpar=list(pch=16, cex=0.7))
plot(grid2,add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('c. Effort correlated, ', hat(italic(D)), ~~'biased')), cex=0.9, col='red')

plot(mask, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(pop, pch=16, cex=0.7, col = 'white')
plot(grid1, add=T, detpar=list(pch=16, cex=0.7))
plot(grid2, add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('d. Detection constant, ', hat(italic(D)), ~~'unbiased')), cex=0.9)

plot(mask, cov='gridy', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(popy, pch=16, cex=0.7, col = 'white')
plot(grid1,add=T, detpar=list(pch=16, fg='orange', cex=0.7))
plot(grid2,add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('e. Detection uncorrelated, ', hat(italic(D)), ~~'biased')), cex=0.9, col='red')

plot(mask, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(pop, pch=16, cex=0.7, col = 'white')
plot(grid1,add=T, detpar=list(pch=16, fg='orange', cex=0.7))
plot(grid2,add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('f. Detection correlated, ', hat(italic(D)), ~~'biased')), cex=0.9, col='red')
```

### Define STR scenarios

#### Varying effort

```{r STR scenarios 1, eval = TRUE, message = FALSE, results = "hide"}
# a. equal effort (2 64-detector grids)
# b. effort uncorrelated: patchy effort unrelated to density  (36-detector, 64-detector grids)
# c. naive: more effort where greater density  (36-detector, 64-detector grids)
# c1. modelling D  (36-detector, 64-detector grids)
# c2. independent estimates  (36-detector, 64-detector grids)

scen1 <- make.scenarios(trapsindex = c(1,2,2,2,2), D = 12.5, detectfn='HHN', 
                       lambda0 = 0.1, sigma = 20, noccasions = 10)
scen1$popindex <- c(1,2,1,1,1)
scen1$fitindex <- c(1,1,1,2,3)
popnset <- list(popargs, popargsy)
trapset1 <- list(grids, gridsa)
fitarg1 <- list(list(detectfn = 'HHN', start = list(D = 12.5)),
               list(detectfn = 'HHN', start = list(D = 12.5), model = D~grid),
               list(detectfn = 'HHN', start = list(D = 12.5),
                    model = list(D~grid, lambda0~grid, sigma~grid)))
```

#### Varying detection

Use 'groups' to define sub-scenarios with differing detection on different subgrids.

```{r STR scenarios 2, eval = TRUE, message = FALSE, results = "hide"}
# d. equal detection (repeats a. so not simulated)
# e. detection uncorrelated: patchy detection unrelated to density
# f. detection correlated
# f1. detection correlated and modelled
# f2. detection correlated, D modelled
# f3. detection correlated and modelled with D

scen2 <- make.scenarios(fitindex=c(1,1,1,2,3,4), D = 12.5, groups = 1:2, detectfn='HHN', 
                        lambda0 = 0.1, sigma = 20, noccasions = 10)
scen2$lambda0[3:12] <- rep(c(0.15, 0.05),5)
scen2$trapsindex <- scen2$group
scen2
popnset <- list(popargs, popargsy)
scen2$popindex <- rep(c(1,2,1,1,1,1), each = 2)
scen2$fitindex <- rep(c(1,1,1,2,3,4), each = 2)
trapset2 <- list(grid1, grid2)
fitarg2 <- list(list(detectfn = 'HHN', start = list(D = 12.5)),
                list(detectfn = 'HHN', start = list(D = 12.5), model = lambda0~grid),
                list(detectfn = 'HHN', start = list(D = 12.5), model = D~grid),
                list(detectfn = 'HHN', start = list(D = 12.5), model = list(D~grid, lambda0~grid))                )
```

```{r STR scenario summary, cache = TRUE}
scenarioSummary(scen1, trapset1)
```

### Run STR simulations

```{r STR run1abc, eval = runsim, cache = TRUE, message = FALSE}
nrepl <- 100
sims1abc <- run.scenarios(nrepl, scen1[1:3,], trapset = trapset1, maskset = mask, 
                      pop.args = popnset, fit = TRUE, fit.args = fitarg1)
saveRDS(sims1abc, file = 'STRsims1abc.RDS')
```

Models with stratum-specific density estimates use `stratumTable` as the extractfn and cov = 'grid' as the stratification mask covariate (which happens in this case to be equal-area as in the default).

```{r STR run1c1c2, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims1c1c2 <- run.scenarios(nrepl, scen1[4:5,], trapset = trapset1, maskset = mask, 
                      pop.args = popnset, fit = TRUE, fit.args = fitarg1,
                      extractfn = stratumTable, cov = 'grid')
saveRDS(sims1c1c2, file = 'STRsims1c1c2.RDS')
```

Simulations with distinct lambda0 on each grid use append.capthist (new in secr 4.6.7) to combine the separately simulated capthist objects before a model is fitted.

```{r STR run2e, eval = runsim, cache = TRUE, message = FALSE}
sims2e <- run.scenarios(nrepl, scen2[3:4,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE)
saveRDS(sims2e, file = 'STRsims2e.RDS')
```

```{r STR run2f, eval = runsim, cache = TRUE, message = FALSE}
sims2f <- run.scenarios(nrepl, scen2[5:6,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE)
saveRDS(sims2f, file = 'STRsims2f.RDS')
```

```{r STR run2f1, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims2f1 <- run.scenarios(nrepl, scen2[7:8,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE)
saveRDS(sims2f1, file = 'STRsims2f1.RDS')
```

```{r STR run2f2, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims2f2 <- run.scenarios(nrepl, scen2[9:10,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE, extractfn = stratumTable, cov='grid')
saveRDS(sims2f2, file = 'STRsims2f2.RDS')
```

```{r STR run2f3, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims2f3 <- run.scenarios(nrepl, scen2[11:12,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE, extractfn = stratumTable, cov = 'grid')
saveRDS(sims2f3, file = 'STRsims2f3.RDS')
```

## Results

`stratumSummary` is 'stratumTable.R'.


```{r STR results, eval = TRUE, echo = TRUE}
STRsims1abc  <- readRDS(file = 'STRsims1abc.RDS')
STRsims1c1c2 <- readRDS(file = 'STRsims1c1c2.RDS')
STRsims2e    <- readRDS(file = 'STRsims2e.RDS')
STRsims2f    <- readRDS(file = 'STRsims2f.RDS')
STRsims2f1   <- readRDS(file = 'STRsims2f1.RDS')
STRsims2f2   <- readRDS(file = 'STRsims2f2.RDS')
STRsims2f3   <- readRDS(file = 'STRsims2f3.RDS')
```

### Varying effort (grid size)

```{r STR results1, eval = TRUE, echo = TRUE}
source('stratumTable.R')
scen1
estimateSummary(STRsims1abc)
stratumSummary(STRsims1c1c2)
```

### Varying detection (parameter lambda0)

```{r STR results2, eval = TRUE, echo = TRUE}
scen2
rbind(
    estimateSummary(STRsims2e, true = 12.5),
    estimateSummary(STRsims2f, true = 12.5),
    estimateSummary(STRsims2f1, true = 12.5)
)
stratumSummary(STRsims2f2)
stratumSummary(STRsims2f3)
```

Note: rRMSE is the root-mean-square error divided by the true density.

$\mbox{rRMSE}(\hat D) = \frac{1}{D} \sqrt {\frac{1}{N} \sum_{i=1}^N {(D - \hat D_i)^2}}$ 

$\mbox{RB}(\hat D) = \frac{1}{N} \sum_{i=1}^N {(\hat D_i - D)}/D$ 


```{r STR figure, eval = FALSE, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Effect of array size on relative bias (RB) and relative root-mean-square error (rRMSE) of density estimates"}
```

