---
title: "**secr** simulations STR"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Stratification and parameter variation STR

The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

Covariation of density and effort or detection leads to bias if not adequately modelled.

<!-- To render from (linux) command line use full file path e.g., -->
<!-- Rscript -e "rmarkdown::render('/home/effmu21p/STR/secr-simulations-STR.rmd')" -->
<!-- Requires pandoc, see e.g., -->
<!-- http://tpepler.github.io/howto/howto-render-r-markdown-from-command-line.html -->

## Settings

Two square grids of 64 binary proximity detectors at spacing $2\sigma$.  

The same conditions were used for simulation and fitting. 

Base scenario

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| D | 12.5 (5,20) |
| lambda0  | 0.1 (0.05,0.15) |
| sigma    | 20 |
| noccasions | 10 |
| buffer   | $4\sigma$ |
| nx       | 64 |

## Package version, date and platform

```{r versions, results = 'hold', echo = FALSE}
source('../setup.R')
source('stratumTable.R')
runsim <- TRUE
metadata(runsim) # if FALSE picks up saved metadata.RDS
```

## Simulation code

### First construct grids

```{r strgrids}
# start with 8 x 8 grids
grid1 <- grid2 <- make.grid(8,8, spacing = 40, detector = 'proximity')
grid2[,1] <- grid2[,1] + 480  # displace to right

covariates(grid1)$grid <- rep(factor(1, levels = 1:2), nrow(grid1))
covariates(grid2)$grid <- rep(factor(2, levels = 1:2), nrow(grid2))

# grids1a is a smaller grid 6 x 6
grid1a <- subset(grid1, !apply(grid1, 1, function(x) any(x<20 | x>260)))

# combine
grids <- rbind(grid1, grid2)
gridsa <- rbind(grid1a, grid2)

# stratify mask horizontally and vertically
mask <- make.mask(grids, buffer = 100, spacing = 10)
covariates(mask)$grid <- factor(1 + (mask$x>380))
covariates(mask)$D <- c(5,20)[covariates(mask)$grid]
covariates(mask)$gridy <- factor(1 + (mask$y>140))
covariates(mask)$Dy <- c(5,20)[covariates(mask)$gridy]

# pops for illustration
popargs <- list(D = 'D', core = mask, model2D = 'IHP')
pop <- do.call(sim.popn, popargs)
popargsy <- list(D = 'Dy', core = mask, model2D = 'IHP')
popy <- do.call(sim.popn, popargsy)
```


### Plot - as in Efford in prep. Study design

```{r strplot, fig.width=8, fig.height=4}
par(mfrow=c(2,3), mar=c(1,1,2,1))

plot(mask, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(pop, pch=16, cex=0.7, col = 'white')
plot(grid1, add=T, detpar=list(pch=16, cex=0.7))
plot(grid2, add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('a. Effort constant, ', hat(italic(D)), ~~'unbiased')), cex=0.9)

plot(mask, cov='gridy', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(popy, pch=16, cex=0.7, col = 'white')
plot(grid1a,add=T, detpar=list(pch=16, cex=0.7))
plot(grid2,add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('b. Effort uncorrelated, ', hat(italic(D)), ~~'unbiased')), cex=0.9)

plot(mask, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(pop, pch=16, cex=0.7, col = 'white')
plot(grid1a,add=T, detpar=list(pch=16, cex=0.7))
plot(grid2,add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('c. Effort correlated, ', hat(italic(D)), ~~'biased')), cex=0.9, col='red')

plot(mask, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(pop, pch=16, cex=0.7, col = 'white')
plot(grid1, add=T, detpar=list(pch=16, cex=0.7))
plot(grid2, add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('d. Detection constant, ', hat(italic(D)), ~~'unbiased')), cex=0.9)

plot(mask, cov='gridy', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(popy, pch=16, cex=0.7, col = 'white')
plot(grid1,add=T, detpar=list(pch=16, fg='orange', cex=0.7))
plot(grid2,add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('e. Detection uncorrelated, ', hat(italic(D)), ~~'biased')), cex=0.9, col='red')

plot(mask, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
points(pop, pch=16, cex=0.7, col = 'white')
plot(grid1,add=T, detpar=list(pch=16, fg='orange', cex=0.7))
plot(grid2,add=T, detpar=list(pch=16, cex=0.7))
mtext (side=3, adj=0, expression(paste('f. Detection correlated, ', hat(italic(D)), ~~'biased')), cex=0.9, col='red')
```

### Define STR scenarios

#### Varying effort

```{r STR scenarios 1, eval = TRUE, message = FALSE, results = "hide"}
# a. equal effort (2 64-detector grids)
# b. effort uncorrelated: patchy effort unrelated to density  (36-detector, 64-detector grids)
# c. naive: more effort where greater density  (36-detector, 64-detector grids)
# c1. modelling D  (36-detector, 64-detector grids)
# c2. independent estimates  (36-detector, 64-detector grids)

scen1 <- make.scenarios(trapsindex = c(1,2,2,2,2), D = 12.5, detectfn='HHN', 
                       lambda0 = 0.1, sigma = 20, noccasions = 10)
scen1$popindex <- c(1,2,1,1,1)
scen1$fitindex <- c(1,1,1,2,3)
popnset <- list(popargs, popargsy)
trapset1 <- list(grids, gridsa)
fitarg1 <- list(list(detectfn = 'HHN', start = list(D = 12.5)),
               list(detectfn = 'HHN', start = list(D = 12.5), model = D~grid),
               list(detectfn = 'HHN', start = list(D = 12.5),
                    model = list(D~grid, lambda0~grid, sigma~grid)))
```

Also try varying allocation between low- and high-density strata. Number of detectors in each varies from 1 to 15 rows of 8 detectors, while keeping total constant at 128 detectors.  Consider both high effort ('v', 10 occasions) and low effort ('v4', 4 occasions).

```{r STR scenarios 1v, eval = TRUE, message = FALSE, results = "hide"}
# c1v. modelling D  (varying grids)
oney <- function (y = 8) {
    grid1 <- make.grid(8,y, detector = 'proximity', spacing = 40)
    grid2 <- make.grid(8,16-y, detector = 'proximity', spacing = 40)
    grid2[,1] <- grid2[,1] + 480  # displace to right
    rbind(grid1, grid2)
}
trapset1v <- lapply(15:1, oney)
maskv <- make.mask(trapset1v[[15]], buffer = 100, spacing = 10)
covariates(maskv)$grid <- factor(1 + (maskv$x>380))
covariates(maskv)$D <- c(5,20)[covariates(maskv)$grid]
scen1v <- make.scenarios(trapsindex = 1:15, D = 12.5, detectfn='HHN', 
                       lambda0 = 0.1, sigma = 20, noccasions = 10, fitindex = 2)

scen1v4 <- make.scenarios(trapsindex = 1:15, D = 12.5, detectfn='HHN', 
                       lambda0 = 0.1, sigma = 20, noccasions = 4, fitindex = 2)
popargsv <- list(D = 'D', core = maskv, model2D = 'IHP')
```

Alternatively, reduce number of occasions in the low-density stratum and correspondingly increase those in the high-density stratum:

```{r STR scenarios 1vocc, eval = TRUE, message = FALSE, results = "hide"}
# c1vocc. modelling D  (varying usage on grids)
oneocc <- function (occ = 1, sumocc = 10) {
    grid1 <- make.grid(8,8, detector = 'proximity', spacing = 40)
    usage(grid1) <- matrix(0,nrow=64, ncol=sumocc-1)
    grid2 <- grid1
    usage(grid1)[,1:occ] <- 1
    usage(grid2)[,1:(sumocc-occ)] <- 1
    grid2[,1] <- grid2[,1] + 480  # displace to right
    rbind(grid1, grid2)
}
trapset1vocc <- lapply(1:9, oneocc)
# check reciprocal variation in effort
t(sapply(trapset1vocc, function(x) apply(usage(x),2,sum)))
# scenarios for total of 10 occasions
scen1vocc <- make.scenarios(trapsindex = 1:9, D = 12.5, detectfn='HHN', 
                       lambda0 = 0.1, sigma = 20, noccasions = 9, fitindex = 2)
```

#### Varying detection

Use 'groups' to define sub-scenarios with differing detection on different subgrids.

```{r STR scenarios 2, eval = TRUE, message = FALSE, results = "hide"}
# d.  equal detection (repeats a. so not simulated)
# e.  detection uncorrelated: patchy detection unrelated to density
# e1. detection uncorrelated and modelled
# f.  detection correlated
# f1. detection correlated and modelled
# f2. detection correlated, D modelled
# f3. detection correlated and modelled with D

scen2 <- make.scenarios(fitindex=c(1,1,2,1,2,3,4), D = 12.5, groups = 1:2, detectfn='HHN', 
                        lambda0 = 0.1, sigma = 20, noccasions = 10)
scen2$lambda0[3:14] <- rep(c(0.15, 0.05),6)
scen2$trapsindex <- scen2$group
popnset <- list(popargs, popargsy)
scen2$popindex <- rep(c(1,2,2,1,1,1,1), each = 2)
trapset2 <- list(grid1, grid2)
fitarg2 <- list(list(detectfn = 'HHN', start = list(D = 12.5)),
                list(detectfn = 'HHN', start = list(D = 12.5), model = lambda0~grid),
                list(detectfn = 'HHN', start = list(D = 12.5), model = D~grid),
                list(detectfn = 'HHN', start = list(D = 12.5), model = list(D~grid, lambda0~grid))                )
```

```{r STR scenario summary, cache = TRUE}
scenarioSummary(scen1, trapset1)
```

### Run STR simulations

```{r STR run1abc, eval = runsim, cache = TRUE, message = FALSE}
nrepl <- 100
sims1abc <- run.scenarios(nrepl, scen1[1:3,], trapset = trapset1, maskset = mask, 
                      pop.args = popnset, fit = TRUE, fit.args = fitarg1)
saveRDS(sims1abc, file = 'STRsims1abc.RDS')
```

Models with stratum-specific density estimates use `stratumTable` as the extractfn and cov = 'grid' as the stratification mask covariate (which happens in this case to be equal-area as in the default).

```{r STR run1c1c2, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims1c1c2 <- run.scenarios(nrepl, scen1[4:5,], trapset = trapset1, maskset = mask, 
                      pop.args = popnset, fit = TRUE, fit.args = fitarg1,
                      extractfn = stratumTable, cov = 'grid')
saveRDS(sims1c1c2, file = 'STRsims1c1c2.RDS')
```

```{r STR run1cv, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims1cv <- run.scenarios(nrepl, scen1v, trapset = trapset1v, maskset = maskv, 
                      pop.args = popargsv, fit = TRUE, fit.args = fitarg1,
                      extractfn = stratumTable, cov = 'grid')
saveRDS(sims1cv, file = 'STRsims1cv.RDS')
```

```{r STR run1cv4, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims1cv4 <- run.scenarios(nrepl, scen1v4, trapset = trapset1v, maskset = maskv, 
                      pop.args = popargsv, fit = TRUE, fit.args = fitarg1,
                      extractfn = stratumTable, cov = 'grid')
saveRDS(sims1cv4, file = 'STRsims1cv4.RDS')
```

```{r STR run1cvocc, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims1cvocc <- run.scenarios(nrepl, scen1vocc, trapset = trapset1vocc, maskset = maskv, 
                      pop.args = popargsv, fit = TRUE, fit.args = fitarg1,
                      extractfn = stratumTable, cov = 'grid')
saveRDS(sims1cvocc, file = 'STRsims1cvocc.RDS')
```


Simulations with distinct lambda0 on each grid use append.capthist (new in secr 4.6.7) to combine the separately simulated capthist objects before a model is fitted.

```{r STR run2e, eval = runsim, cache = TRUE, message = FALSE}
sims2e <- run.scenarios(nrepl, scen2[3:4,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE)
saveRDS(sims2e, file = 'STRsims2e.RDS')
```

```{r STR run2e1, eval = runsim, cache = TRUE, message = FALSE}
sims2e1 <- run.scenarios(nrepl, scen2[5:6,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE)
saveRDS(sims2e1, file = 'STRsims2e1.RDS')
```

```{r STR run2f, eval = runsim, cache = TRUE, message = FALSE}
sims2f <- run.scenarios(nrepl, scen2[7:8,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE)
saveRDS(sims2f, file = 'STRsims2f.RDS')
```

```{r STR run2f1, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims2f1 <- run.scenarios(nrepl, scen2[9:10,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE)
saveRDS(sims2f1, file = 'STRsims2f1.RDS')
```

```{r STR run2f2, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims2f2 <- run.scenarios(nrepl, scen2[11:12,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE, extractfn = stratumTable, cov='grid')
saveRDS(sims2f2, file = 'STRsims2f2.RDS')
```

```{r STR run2f3, eval = runsim, cache = TRUE, warning = FALSE, message = FALSE}
sims2f3 <- run.scenarios(nrepl, scen2[13:14,], trapset = list(grid1,grid2), maskset = mask, 
                       pop.args = popnset, multisession = append.capthist,
                       fit.args = fitarg2, fit = TRUE, extractfn = stratumTable, cov = 'grid')
saveRDS(sims2f3, file = 'STRsims2f3.RDS')
```

## Results

Effort scenarios involve either two equal 64-detector grids or one 36-detector grid and one 64-detector grid.

Detection scenarios involve two equal 64-detector grids, with equal $\lambda_0$ (0.1) or unequal $\lambda_0$ (0.15,0.05).

| Code | Series | Scenario | Description | Model | RB($\hat D$) |
|:---|:---|:---|:------------|:-------------|:-------|
| a  | 1 | 1 | equal effort | D~1, lambda0~1, sigma~1 | +0.02 |
| b  | 1 | 2 | effort uncorrelated | D~1, lambda0~1, sigma~1 | +0.02 | 
| c  | 1 | 3 | effort high where greater density | D~1, lambda0~1, sigma~1 | **+0.17** |
| c1 | 1 | 4 | | D~grid, lambda0~1, sigma~1 | 0.00* |
| c2 | 1 | 5 | | D~grid, lambda0~grid, sigma~grid | +0.02* |
| d  | 2 | 1 | equal detection (repeats a) | D~1, lambda0~1, sigma~1 |  |
| e  | 2 | 2 | detection uncorrelated | D~1, lambda0~1, sigma~1 | **--0.14**|
| e1 | 2 | 3 |  | D~1, lambda0~grid, sigma~1 | 0.00 |
| f  | 2 | 4 | detection high where low density | D~1, lambda0~1, sigma~1 | **--0.15** |
| f1 | 2 | 5 |  | D~1, lambda0~grid, sigma~1 | **--0.18** |
| f2 | 2 | 6 |  | D~grid, lambda0~1, sigma~1 | **--0.15** * |
| f3 | 2 | 7 |  | D~grid, lambda0~grid, sigma~1 | +0.01 * |

Table: Summary of simulations. Large relative bias in **bold** face.
\* area-weighted estimate of total density from stratified model

```{r STR results, eval = TRUE, echo = TRUE}
STRsims1abc  <- readRDS(file = 'STRsims1abc.RDS')
STRsims1c1c2 <- readRDS(file = 'STRsims1c1c2.RDS')
STRsims1cv   <- readRDS(file = 'STRsims1cv.RDS')
STRsims1cv4  <- readRDS(file = 'STRsims1cv4.RDS')
STRsims1cvocc <- readRDS(file = 'STRsims1cvocc.RDS')
STRsims2e    <- readRDS(file = 'STRsims2e.RDS')
STRsims2e1   <- readRDS(file = 'STRsims2e1.RDS')
STRsims2f    <- readRDS(file = 'STRsims2f.RDS')
STRsims2f1   <- readRDS(file = 'STRsims2f1.RDS')
STRsims2f2   <- readRDS(file = 'STRsims2f2.RDS')
STRsims2f3   <- readRDS(file = 'STRsims2f3.RDS')
```

### Varying effort (grid size)

`stratumSummary` is in 'stratumTable.R'.

```{r STR results1, eval = TRUE, echo = TRUE}
source('stratumTable.R')
scen1[, -c(5,10)]
estimateSummary(STRsims1abc)
stratumSummary(STRsims1c1c2)
stratumSummary(STRsims1cv)
stratumSummary(STRsims1cv4)
stratumSummary(STRsims1cvocc)
```

```{r STR scenarios 1vplot, eval = TRUE, fig.width=8, fig.height=2.5, message = FALSE, results = "hide"}
plotone <- function(tr) {
    plot(maskv, cov='grid', legend = FALSE, dots=F, col=c('lightgreen','lightblue'))
    plot (tr, add = TRUE, detpar=list(pch=16, cex=0.5))
}
par(mfrow = c(1,5), mar = c(1,1,1,1), oma = c(3,1,1,1))
sapply(trapset1v[c(1,4,8,12,15)], plotone)
mtext(paste0(c(1,4,8,12,15), '/16'), outer=TRUE, side = 1, line = -2, 
      at= c(0.1, 0.3, 0.5, 0.7, 0.9), cex = 0.8)
mtext('Fraction of effort in high-density stratum', 
      outer = TRUE, side=1, line=0, cex=0.9)
```

Figure: Varying allocation of effort; examples from full spectrum. 

```{r STR results1v, eval = TRUE, echo = TRUE}
source('stratumTable.R')
ss1cv <- stratumSummary(STRsims1cv)
ss1cv4 <- stratumSummary(STRsims1cv4)
par(mfrow = c(1,1))
plot(0,0, type = 'n', xlim = c(0,1), ylim = c(0.0,0.30), 
     xlab = 'Fraction of effort',
     ylab = 'RSE D-hat')
x <- (1:15)/16
rse1cv <- sapply(ss1cv, '[', 'Total','RSE')
rse1cv4 <- sapply(ss1cv4, '[', 'Total','RSE')
points(x, rse1cv, pch = 16)
points(x, rse1cv4, pch = 1)
```

Figure: Effect of varying allocation of effort on precision of $\hat D$. Open circles 4 occasions, filled circles 10 occasions. 

### Varying detection (parameter lambda0)

```{r STR results2, eval = TRUE, echo = TRUE}
scen2[, -c(5,10)]
rbind(
    estimateSummary(STRsims2e,  true = 12.5),
    estimateSummary(STRsims2e1, true = 12.5),
    estimateSummary(STRsims2f,  true = 12.5),
    estimateSummary(STRsims2f1, true = 12.5)
)
stratumSummary(STRsims2f2)
stratumSummary(STRsims2f3)
```

Note: rRMSE is the root-mean-square error divided by the true density.

$\mbox{rRMSE}(\hat D) = \frac{1}{D} \sqrt {\frac{1}{N} \sum_{i=1}^N {(D - \hat D_i)^2}}$ 

$\mbox{RB}(\hat D) = \frac{1}{N} \sum_{i=1}^N {(\hat D_i - D)}/D$ 


```{r STR figure, eval = FALSE, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Effect of array size on relative bias (RB) and relative root-mean-square error (rRMSE) of density estimates"}
```

