---
title: "**secr** simulations TEL"
author: "Murray Efford and John Boulanger"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Telemetry marking

These simulations test a MLE implementation of the 1-step method of Whittington et al. (2025a).

## Problem description

It has been generally recognised that spatial mark--resight studies require either data on the initial deployment of marks, allowing the spatial distribution of marked animals to be modelled, or strong assumptions about the distribution (e.g. spatially uniform) (Royle et al. 2014, Efford and Hunter 2018).

However, the papers of Margenau et al. (2022) and Whittington et al. (2025a) pointed to a way of analysing camera data for collared animals that is not prone to bias from unmodelled deployment of collars. They recognised that detection parameters (e.g. $\lambda_0$, $\sigma$) may be estimated from the collared animals alone. Combining the detection parameter estimates with observed counts in either a 2-stage- (Margenau et al. 2022) or 1-stage-process (Whittington et al. 2025a) yields estimates of population density.

We consider the following scenario

* data include telemetry locations for a sample of animals,
* deployment (attachment) of telemetry devices (e.g., radiocollars) is not modelled,
* the spatial distribution of AC of telemetered animals is unknown,
* detection histories (binary proximity or count) relate only to pre-marked telemetered individuals, 
* there is a history for each telemetered individual (all-zero if not detected), and
* equivalent data (counts of detections) are collected for all individuals, whether marked and unmarked.

## Implementation in **secr** 5.5.0

This is not described here in detail. Basically, 

1. A capthist object is formed from the sightings of telemetered animals using 
the new telemetry type 'marking', the markocc attribute of its 'traps' is set to 
0 for sighting occasions and 1 for the telemetry occasion(s), and sightings of 
unmarked animals are attached. The required capthist has one row for each 
telemetered animal, a 'telemetryxy' attribute that is a list with telemetry 
locations, and attribute 'Tu' for additional sightings of unmarked animals.
2. A model is fitted as usual. The new 'details' settings 'uselog' and 'safeLL' 
should be TRUE to avoid numerical issues.
3. At the same time or later, sighting data are simulated to determine the 
overdispersion 'chat' and adjust the variances (Efford and Hunter 2018).

## Settings

These are intended to follow the 24 scenarios considered by Whittington et al. (2025a Appendix Table S2).  A grid of 81 cameras spanning 90 units on a side (Whittington et al. 2025a p.4 stated 120 units) lies within a 150 x 150 square region (possibly 149 x 149 in Whittington et al. 2025b).

The same conditions used for simulation and fitting.

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| N        | 60, 120 |    
| D | c(60,120) / ca2.25 ha |
| lambda0  | low or high (each varies with sigma) |
| sigma    | 4 or 8 units |
| noccasions | 5 |
| buffer   | 30 units |
| mask spacing | 2.5 or 5 units (0.625 sigma) |

Four combinations of lambda0 and sigma are used (see scen).

## Package version, date and platform

```{r startup, warning = FALSE, message = FALSE, results = "hold"}
# source('../setup.R')
source('d:/density secr 5.5/knitr/secr-simulations/setup.R')
runsim  <- FALSE
figures <- TRUE
cache   <- FALSE
nrepl   <- 200
metadata(runsim) # if FALSE picks up saved metadata.RDS
```

## Simulation code

The simulation function `fnc.create.SMR.data` from file '1a._gSMR_simStudy_functions_and_nimble_code.R' of Whittington et al. (2025b) is used to generate each replicate capthist. It is copied here, along with the required function `e2dist`.

```{r Whittington functions}
# See Whittington et al. (2025b)

# 0. Function e2dist ####
e2dist <- function (x, y) {  # Function from scrbook package to calculate the distance between locations in 2 matrices.
  i <- sort(rep(1:nrow(y), nrow(x)))
  dvec <- sqrt((x[, 1] - y[i, 1])^2 + (x[, 2] - y[i, 2])^2)
  matrix(dvec, nrow = nrow(x), ncol = nrow(y), byrow = F)
}

# 0.  Function to simulate data ####

fnc.create.SMR.data <- function (N = N, nMarked = nMarked, sigma = 1, lam0 = 0.5, K = 5,
                                 n.camera.row = 9, 
                                 xlims = c(1, 150), ylims = c(1, 150), xlim.R = c(30, 120), ylim.R = c(30, 120),
                                 nlocs = 100, plotSim = TRUE) 
  # N = True number of animals in the state-space
  # nMarked = number of animals to capture and mark
  # sigma = scale parameter for home range size
  # lam0 = encounter rate at home range centre for resighting
  # K = number of resighting (eg camera) occasions
  # n.camera.row = sqrt(total number of cameras)
  # xlims = min and max coordinates of easting for the state-space
  # ylims = min and max coordinates of northing for the state-space
  # xlim.R = min and max coorindates of easting for the study area
  # ylim.R = min and max coordinates of northing for the study area.
  # nMarked = Is the number of marked animals.  
  # nlocs = number of telemetry locations for each marked animal.  
  # plotSim = whether to plot state space, detectors, capture summary, and resighting summary
{
  ## Camera coordinates
  coor0 <- seq(xlim.R[1], xlim.R[2], length = n.camera.row)
  X.camera <- cbind(rep(coor0, each=length(coor0)),rep(coor0, times=length(coor0))) 
  
  J.camera <- nrow(X.camera)  
  # Capture Process: activity centres and marked animals
  enoughMarked = FALSE
  counter <- 0
  while(enoughMarked == FALSE & counter < 100){
    counter <- counter + 1
    sx <- runif(N, xlims[1], xlims[2])
    sy <- runif(N, ylims[1], ylims[2])
    sxy <- round(cbind(sx, sy))

    #### Capture data
    D <- e2dist(sxy, X.camera)
    
    K_capt = 20  #K
    lam_c <- lam0 * exp(-(D * D)/(2 * sigma * sigma))
    Y_c <- array(NA, c(N, J.camera, K_capt))   
    for (i in 1:N) {
      for (j in 1:J.camera) {
        Y_c[i, j, ] <- rpois(K_capt, lam_c[i, j]) # Detections of each individual
      }
    }
    # Select the first nMarked individuals captured
    for (k in 1:K_capt){
      marked.k <- apply(Y_c[ , , 1:k], 1, sum) 
      marked.01.k <- marked.k > 0 * 1
      if (sum(marked.01.k) >= nMarked){
        Y_c[ , , (k + 1):K_capt] <- 0
        break()
      }
    }
    indiv.captures <- apply(Y_c, 1, sum)              
    indiv.capt01 <- (indiv.captures > 0) * 1
    iMarked <- which(indiv.capt01 == 1)
    nDetected = length(iMarked)
    if(nDetected >= nMarked){
      enoughMarked = TRUE
    }
  } # End While loop
  if (enoughMarked == FALSE){
    stop('... change parameters - not enough animals within the study area for nMarked')
  }
  # select marked animals
  iKeep <- sample(1:N, size = nMarked, replace = FALSE, prob = indiv.captures / sum(indiv.captures))
  # Re-arrange so activity centres ordered by marked, then unmarked
  tmp1 <- sxy[iKeep, ]      # marked individuals
  tmp2 <- sxy[-c(iKeep), ]  # unmarked individuals
  # Arrange animals by marked, then unmarked
  sxy <- rbind(tmp1, tmp2)
  marked = ifelse(1:N <= nMarked, 1, 0)
  #  Resight data
  D <- e2dist(sxy, X.camera)
  lam <- lam0 * exp(-(D * D)/(2 * sigma * sigma))
  Y <- array(NA, c(N, J.camera, K))
  for (i in 1:N) {
    for (j in 1:J.camera) {
      Y[i, j, ] <- rpois(K, lam[i, j]) # true detections of each individual
    }
  }
  indiv.detections <- apply(Y, 1, sum)             
  indiv.detected <- (indiv.detections > 0) * 1
  nDet.jk <- apply(Y, c(2, 3), sum)   #  Sum of detections by Camera and Occasion
  det01.jk <- (nDet.jk > 0) * 1       #  Binomial detections
  nDet_population <- rowSums(nDet.jk)
  y01_population <- rowSums(det01.jk)
  Y.marked <- Y[marked == 1, , ]
  Y.marked.01 <- (Y.marked > 0) * 1  # i, j, k
  nDet_marked <- apply(Y.marked, c(1, 2), sum)  # individual x trap
  y01_marked <- apply(Y.marked.01, c(1, 2), sum) # matrix i x j for number of occasions with a detection

  telem_array <- array(0, dim = c(nMarked, nlocs, 2))
  if (nlocs > 0) {
    locs <- list()
    for (i in 1:nMarked) {
      telem_array[i, 1:nlocs, 1] <- rnorm(nlocs, sxy[i, 1], sigma)
      telem_array[i, 1:nlocs, 2] <- rnorm(nlocs, sxy[i, 2], sigma)
      
    }
  }
  if (plotSim == TRUE){
    par(mfrow = c(1,1))
    plot(sxy, col = 'red', pch = 19, xlim = xlims, ylim = ylims, asp = 1, xlab = 'X', ylab = 'Y',
         main = paste('Spatial Mark-Resight Simulated Data \n ',
                      ' lam0 =', lam0, ' sigma =', sigma, ' N =', N, sep = ' '))
    rect(xleft = xlims[1], xright = xlims[2], ybottom = ylims[1], ytop = ylims[2], col = 'gray80' )
    rect(xleft = xlim.R[1], xright = xlim.R[2], ybottom = ylim.R[1], ytop = ylim.R[2], col = 'gray70' )
    lapply(seq(xlims[1], xlims[2], by = 10), function(x) lines(c(x, x), c(ylims[1], ylims[2]), col = 'white'))
    lapply(seq(ylims[1], ylims[2], by = 10), function(x) lines( c(xlims[1], xlims[2]), c(x, x), col = 'white'))
    for(ttt in 1:nMarked){
      points(telem_array[ttt, ,1:2], col = ttt, cex = 0.4, pch = 1)
    }
    points(sxy, col = 'red', pch = 19, cex = 1.3)
    #points(X.trap, col='blue', pch = 1, cex = 1.2)
    points(X.camera, col = 'forestgreen', pch = '+', cex = 0.8)
    points(sxy[marked == 1, 1], sxy[marked == 1, 2], pch=19, col='purple', cex = 1.3)
    points(sxy[indiv.detected == 1, 1], sxy[indiv.detected == 1, 2], pch=1, col='yellow', cex = 0.5)
    legend(x = 0.7 * xlims[2], y = ylims[2], cex = 0.7, legend = c(  'Cameras', 'Home Range Centre', '  Marked', '  Detected Camera'),
           pch = c( 3, 19, 19, 1), col = c('forestgreen', 'red','purple', 'yellow' ), bg = 'gray80')
  }
  
  list(
      N               = N, 
      nMarked         = nMarked, 
      sigma           = sigma, 
      lam0            = lam0,
      sxy             = sxy, 
      Y               = Y,
      nDet_population = nDet_population,
      nDet_marked     = nDet_marked,
      y01_population  = y01_population,
      y01_marked      = y01_marked,
      telem_array     = telem_array, 
      X.camera        = X.camera,
      prop.detected   = mean(indiv.detected)
  )
  
}   # END OF FUNCTION  fnc.create.SMR.data
```

To use this function in **secrdesign** we define two further functions, one to cast 
the output from `fnc.create.SMR.data` as a capthist object and one to call the simulation 
function in a way that suits the 'CH.function' argument of `secrdesign::run.scenarios`.

```{r wrapper functions}
# cast output from fnc.create.SMR.data as capthist object
asCH <- function(Wdata, detector = 'count', verify = FALSE) {
    with (Wdata, {
        nlocs <- dim(telem_array)[2]
        dimnames(X.camera)[[2]] <- c('x','y')
        trdf <- as.data.frame(X.camera)
        ch <- aperm(Y[1:nMarked,,,drop = FALSE], c(1,3,2))
        class(ch) <- 'capthist'
        traps(ch) <- read.traps(data = trdf, detector = detector)
        rownames(ch) <- 1:nMarked
        tu <- apply(Y[(nMarked+1):N,,], 2:3, sum, drop = FALSE)
        telemID <- data.frame(Session = rep(1,nMarked*nlocs), 
                              ID = rep(1:nMarked, each = nlocs),
                              Occasion = rep(1:nlocs, nMarked))
        telemmat <- matrix((aperm(telem_array, c(2,1,3))), ncol = 2)
        telemdf <- as.data.frame(telemmat)
        te <- read.telemetry(data = cbind(telemID, telemdf), verify = FALSE)
        ch <- addTelemetry(ch, te, type = 'marking', verify = FALSE)
        markocc(traps(ch)) <- c(rep(0,ncol(tu)),1)
        addSightings(ch, unmarked = tu, verify = verify)
    })
}
# function to call fnc.create.SMR.data and asCH from within run.scenarios
# this is an analogue of secr::sim.capthist
getWCH <- function (traps, popn, detectfn, detectpar, noccasions, 
                      ntelem, nocctelem, plotSim = FALSE) {
    # form capthist object from secrdesign 'CH.function' inputs 
    # but using Whittington et al. 2025b simulation code
    if (detectfn != 14 && detectfn != 'HHN') stop ("only for detectfn HHN")
    N            <- nrow(popn)
    nMarked      <- ntelem
    K.camera     <- noccasions
    sigma        <- detectpar$sigma
    lam0         <- detectpar$lambda0
    n.camera.row <- sqrt(nrow(traps))
    xlims        <- range(attr(popn, 'boundingbox')$x)
    ylims        <- range(attr(popn, 'boundingbox')$y)
    xlim.R       <- range(traps$x)
    ylim.R       <- range(traps$y)
    nlocs        <- nocctelem
    Wdata <- fnc.create.SMR.data(
        N = N, nMarked = nMarked,  K = K.camera, sigma = sigma, lam0 = lam0, 
        n.camera.row = n.camera.row, 
        xlims = xlims, ylims = ylims, xlim.R = xlim.R, ylim.R = ylim.R,  
        nlocs = 100, plotSim = plotSim) 
    asCH (Wdata, detector = 'count', verify = FALSE)
}
```
### Define TEL scenarios for secrdesign

```{r TEL scenarios, eval = TRUE, message = FALSE, results = "hide"}
trW        <- make.grid(detector = "count", nx = 9, ny = 9, spacing = 90/8, 
                        origin = c(30,30))

# Define habitat masks with discretization appropriate to sigma = 4, sigma = 8
# The chosen spacing (0.625 sigma) is probably safe for unbiased estimation of density
# (Efford 2025 Fig. 12.5) but may not be perfect for lambda0 and sigma
mskW25     <- make.mask (trW, buffer = 30, type = 'traprect', spacing = 2.5)
mskW50     <- make.mask (trW, buffer = 30, type = 'traprect', spacing = 5)

# scenario defines abundance in terms of density (animals/ha) instead of N
scen <- make.scenarios(noccasions = 5,   # number of sighting occasions 
                       D          = c(60,120)/1.5^2,   # but (1-150) x (1-150) in original? 
                       lambda0    = c(0.62,0.18,1.85,0.54), 
                       sigma      = 4,      # sigma = 8 conditional on lambda0 below
                       detectfn   = 'HHN',  # hazard halfnormal, detectfn = 14
                       detindex   = 1:3)
scen$sigma[scen$lambda0 %in% c(0.18, 0.54)] <- 8
# sort rows to match Whittington et al.
scen <- scen[order(scen$D, scen$sigma, scen$lambda0, scen$detindex),]
# reassign scenario numbers
rownames(scen) <- scen$scenario <- 1:nrow(scen)   
scen$maskindex <- scen$sigma/4   # 1 for mskW25, 2 for mskW50
scen
```

```{r TEL otherargs}
poparg <- list(Ndist = 'fixed')
detarg <- list(list(nocctelem = 100, ntelem = 5),
               list(nocctelem = 100, ntelem = 10),
               list(nocctelem = 100, ntelem = 20))
fitarg <- list(detectfn = 'HHN', 
               details  = list(
                   distribution = 'binomial',    # because N fixed, not Poisson
                   uselog = TRUE, 
                   safeLL = TRUE))
```


### Illustrate one simulation

```{r testing, eval = TRUE, echo = TRUE, fig.width = 6, fig.height = 6, out.width = '80%'}
par(mar = c(3,3,3,2))
popn <- sim.popn(D = 60/2.25, trW, buffer = 30, Ndist = 'fixed')
ch <- getWCH (trW, 
              popn       = popn,
              detectfn   = 'HHN', 
              detectpar  = list(lambda0 = 0.62, sigma = 4), 
              noccasions = 5, 
              ntelem     = 10,
              nocctelem  = 100,
              plotSim    = TRUE)
# and print summary
summary(ch)
```

### Run TEL simulations

```{r TEL run, eval = runsim, cache = cache, warning = FALSE, message = FALSE}
TELsims <- run.scenarios(
    nrepl       = nrepl, 
    scenarios   = scen, 
    CH.function = "getWCH",
    trapset     = list(trW), 
    maskset     = list(mskW25, mskW50), 
    pop.args    = poparg,
    det.args    = detarg, 
    fit         = TRUE, 
    fit.args    = fitarg, 
    seed        = 123,
    chatnsim    = -10000,    # simulations to adjust variance 
    verbose     = FALSE)
saveRDS(TELsims, file = 'TELsims.RDS')
```

```{r TEL readRDS, echo = FALSE}
# load simulations, possibly from a previous run (runsim = FALSE)
TELsims <- readRDS(file = 'TELsims.RDS')
```

TELsims completed with `r nrepl` replicates on `r setNumThreads()` threads in 
`r round(TELsims$proctime/60,1)` minutes.

## Results

### Timing

'proctime' is the time in seconds for a single replicate.  
'chat' is the overdispersion factor used to correct the variances from the independent-Poisson model of sightings (to allow for aggregation across cameras and occasions).

These are averages.

```{r proctime}
TELattr <- data.frame(
    proctime = apply(sapply(TELsims$output, function(x) sapply(x, attr, 'proctime')),2,mean, na.rm = TRUE),
    chat = apply(sapply(TELsims$output, function(x) sapply(x, attr, 'chat')),2,mean, na.rm = TRUE)
)
TELattr
```

```{r altproctime, eval = FALSE, echo = FALSE}
# alternative summary of proctime - not used
addAttributes <- function (object) {
    oneTime <- function(x) {
        time <- attr(x,'proctime')
        chat <- if (is.null(attr(x,'chat'))) NA else attr(x,'chat')[1]
        newdf <- data.frame(
            link=c('',''), estimate = c(time, chat), SE.estimate = c(0,0), 
            lcl = c(NA,NA), ucl = c(NA,NA), row.names = c("proctime","chat"))
        rbind(x, newdf)
    }
    object$output <- lapply(object$output, function (y) lapply(y, oneTime))
    object
}
tmp <- addAttributes(TELsims)
estimateSummary(tmp, parameter = 'proctime', statistics = c('nvalid', 'EST', 'seEST'))
```

```{r TEL tables, eval = figures, echo = FALSE}
# before printing tables of estimates...
options(width = 110, digits = 4)
estD <- estimateSummary(TELsims)[,-1]
estL <- estimateSummary(TELsims, 'lambda0')[,-1]
estS <- estimateSummary(TELsims, 'sigma')[,-1]
scenarios <- scen[,c(5,6,7)]
```

<!-- ### Raw counts -->

<!-- ```{r TEL tables2, eval = figures, echo = FALSE, results = "hold"} -->
<!-- cbind(scenarios, round(getcounts(TELsims),4)) -->
<!-- ``` -->

### Estimates

```{r TEL tables3, eval = figures, echo = FALSE, results = "hold"}
cat('Density\n')
cbind(scenarios, round(estD,4))

cat('\nlambda0\n') 
cbind(scenarios, round(estL,4))

cat('\nsigma\n')
cbind(scenarios, round(estS,4))

cat('\nNote biases in lambda0 and sigma are compensatory\n')
```
 
### Compare estimated and empirical variance of density estimate
 
The estimated SE should match the SD of the simulated estimates.

```{r moreresults, eval = figures}
par(pty = 's')
plot(apply(estimateArray(TELsims)[1,1,1,,],1,sd, na.rm = TRUE),
     apply(estimateArray(TELsims)[1,2,1,,],1,mean, na.rm = TRUE),
     xlab = 'Empirical SE(D)', ylab = 'Estimated SE(D)', xlim = c(0,25), ylim = c(0,25))
abline(0,1)
```
 
## Comparisons and Interpretation

Why might results differ from Appendix Table S2 of Whittington et al. (2025a) 
(ecs270246-sup-0001-appendixs1.pdf)?

```{r S2, eval = TRUE, mar=c(3,3,1,1)}
# read text file containing rows of Whittington et al. 2025a Appendix Table S2,
# excluding those for 2-phase analyses
S2 <- read.table ('AppTableS2.txt', header = TRUE)
print(S2, row.names = FALSE)
```

A key difference is that their algorithm set maximum estimates of D/0.6 i.e. {r round(c(60,120)/2.25/0.6, 2)}. We can apply that filter to the current results:

```{r simsum, eval = TRUE, mar=c(3,3,1,1)}
# read simulation results
TELsims <- readRDS('TELsims.RDS')
# define matrix with min,max density for each scenario
permitted <- outer(TELsims$scenarios$D, c(0, 1/0.6)) 
# summarise filtered results
simsum <- estimateSummary(TELsims, 'D', validrange = permitted)
simsum
```

Comparing current simulations ('current') with Whittington et al. (2025a) ('Table S2') --

```{r simsum figures, eval = TRUE, fig.width=7, fig.height=7.5, out.width = '100%'}
par(pty = 's', mfrow = c(2,2), mar = c(3,3,3,2), mgp = c(2,0.7,0))

plot(S2$Bias, simsum$RB*100, xlab = 'RB Table S2', ylab = 'RB current',
     xlim=c(-10,12), ylim=c(-10,12))
text(S2$Bias+0.6, simsum$RB*100, 1:24, cex = 0.6)
abline(h=0, v=0)
mtext(side=3, line=0.5, 'Bias(D-hat) %', cex=0.9)

plot(S2$CVmean, simsum$RSE*100, xlab = 'CVmean Table S2', ylab = 'RSE current',
     xlim=c(0,50), ylim=c(0,50))
text(S2$CVmean+2, simsum$RSE*100, 1:24, cex=0.6)
abline(0,1)
mtext(side=3, line=0.5, 'RSE(D-hat) %', cex=0.9)

plot(S2$RMSE, simsum$RMSE, xlab = 'RMSE Table S2', ylab = 'RMSE current',
     xlim=c(0,25), ylim=c(0,25))
text(S2$RMSE+0.8, simsum$RMSE, 1:24, cex=0.6)
abline(0,1)
mtext(side=3, line=0.5, 'RMSE(D-hat)', cex=0.9)

plot(S2$CIC, simsum$COV*100, xlab = 'CIC Table S2', ylab = 'COV current',
     xlim=c(88,100), ylim=c(88,100))
text(S2$CIC+0.4, simsum$COV*100, 1:24, cex=0.6)
mtext(side=3, line=0.5, 'Coverage 95% CI for D-hat', cex=0.9)
abline(h=95,v=95)
```

### Notes

* 'RB current' may be inflated by 1 - 149^2^ / 150^2^  = +1.3% if N was distributed over the smaller region 149 x 149.
* RMSE must have been calculated differently ('RMSE Table S2' about twice 'RMSE current')
* Both methods have excessive CI coverage

### Compare timing

Average times in seconds --

```{r, results = 'hold'}
se <- function(x) sd(x, na.rm=TRUE) / sqrt(sum(!is.na(x)))
cat('Current\n')
c(mean = mean(TELattr$proctime), SE = se(TELattr$proctime))
cat('\nTable S2\n')
c(mean = mean(S2$Time*60), SE = se(S2$Time*60))
```

# References

Chandler, R. B., Crawford, D. A., Garrison, E. P., Miller, K. V., and Cherry, M. J. (2021) Modeling abundance, distribution, movement and space use with camera and telemetry data. Ecology 103: e3583. https://doi.org/10.1002/ecy.3583

Efford, M. G. and Hunter, C. M. (2018) Spatial capture--mark--resight estimation of animal population density. Biometrics 74: 411--420. https://doi.org/10.1111/biom.12766

Efford, M. G. and Mowat, G. (2014) Compensatory heterogeneity in spatially explicit capture–recapture data. 
Ecology 95: 1341--1348. https://www.otago.ac.nz/density/pdfs/Efford%20and%20Mowat%202014%20Ecology.pdf

Efford, M. G. (2025) The SECR book. A handbook of spatially explicit capture–recapture methods. Version 1.0.0. Zenodo https://doi.org/10.5281/zenodo.15109938.

Margenau, L. L. S., Cherry, M. J., Miller, K. V., Garrison, E. P., and Chandler, R. B. (2022)
Monitoring partially marked populations using camera and telemetry data. Ecological Applications 32: e2553.
https://doi.org/10.1002/eap.2553

Meyer, C. J., Whittington, J., Tourani, M., and Hebblewhite, M. (2026) A novel generalized spatial mark‐resight model that accounts for group associations. Methods in Ecology and Evolution 17: 1275--1289. https://doi.org/10.1111/2041-210x.70271

Murphy, S. M., Wilckens, D. T., Augustine, B. C., Peyton, M. A., and Harper, G. C. (2019) Improving estimation of puma (*Puma concolor*) population density: clustered camera-trapping,
telemetry data, and generalized spatial mark-resight models. Scientific Reports 9: 4590 https://doi.org/10.1038/s41598-019-40926-7

Royle, J. A., Chandler, R. B., Sollmann, R. and Gardner, B. (2014)
*Spatial capture--recapture.* Academic Press.

Whittington, J., Hebblewhite, M., and Chandler, R. (2018) Generalized spatial mark-resight models with an application to grizzly bears. Journal of Applied Ecology 55: 157--168. https://doi.org/10.1111/1365-2664.12954

Whittington, J., Hebblewhite, M., Meyer, C., Johnston, B., Forshner, A., Macbeth, B. J., Einfeldt, A. L., and Cherry, S. G. (2025a) One‐stage spatial mark–resight analysis reveals an increasing grizzly bear population with declining density near roads. Ecosphere 16: e70246. https://doi.org/10.1002/ecs2.70246

Whittington, J., Hebblewhite, M., Meyer, C., Johnston, B., Forshner, A., Macbeth, B., Einfeldt, A., and Cherry, S. (2025b). Data from: One-stage spatial mark-resight analysis reveals an increasing grizzly bear population with declining density near roads [Computer software]. Zenodo. https://doi.org/10.5281/zenodo.10198936


# Appendix. Alternate CH.function for de novo simulation 

This is code that I used before realising I could use Jesse's code directly.

```{r simulationold}
# Code to simulate combined telemetry-sighting-camera capthist object
Tafunction <-  function (traps, popn, detectfn, detectpar, noccasions, 
                         ntelem, nocctelem, uniformtelem = FALSE, seed = NULL) {
    # select sample of telemetered animals
    if (uniformtelem) {
        telem <- sample.int(nrow(popn), ntelem)
    }
    else {
        # Whittington et al. 2025 selected first ntelem captured animals (cheating?)
        # this is an analogue (inclusion proportional to sighting probability)
        p <- pdot(popn, traps, detectfn, detectpar, noccasions = 1)
        telem <- sample.int(nrow(popn), ntelem, prob = p)
    }
    
    popt  <- subset(popn, telem)
    # others are not telemetered
    popnt <- subset(popn, rownames(popn)[!rownames(popn) %in% telem])
    
    # simulate detections of telemetered animals
    # renumber = FALSE (keep original animalID) needed for matching
    tCH <- sim.capthist(tr,  
                        popn       = popt, 
                        renumber   = FALSE, 
                        detectfn   = detectfn, 
                        noccasions = noccasions,
                        detectpar  = detectpar,
                        seed       = seed)
    
    # simulate sightings of non-telemetered animals
    # these contribute only sightings
    nontCH <- sim.capthist(tr,  
                           popn       = popnt, 
                           renumber   = FALSE, 
                           detectfn   = detectfn, 
                           noccasions = noccasions,
                           detectpar  = detectpar)
    
    # simulate some telemetry
    te <- make.telemetry()
    teCH <- sim.capthist(te, 
                         popn       =  popt, 
                         renumber   = FALSE, 
                         detectfn   = detectfn, 
                         detectpar  = list(lambda0 = 1, sigma = detectpar$sigma),
                         noccasions = nocctelem)
    
    # combine simulated telemetry and sightings of telemetered animals
    # telemetry collapsed to a single occasion
    combinedCH <- addTelemetry(tCH, teCH, type = "marking", collapsetelemetry = TRUE, verify = FALSE)
    markocc(traps(combinedCH)) <- c(rep(0,noccasions),1)
    combinedCH <- addSightings(combinedCH, unmarked = apply(nontCH, c(3,2), sum), verify = FALSE)
    combinedCH
}
```
