---
title: "**secr** simulations OU 2"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Spatial autocorrelation OU Version 2

The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

## Problem description

SECR assumes successive detections of an individual are independent events, conditional on the location of the activity centre. The Ornstein-Uhlenbeck distribution provides a serially correlated model of animal movement that may be used to test the effect of non-independence on SECR estimates.

## Settings

Square grid of 64 count proximity detectors at spacing 100 m.  

The same conditions were used for simulation and fitting. 

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| D | 2 / ha |
| lambda0  | NA |
| sigma    | 100 |
| tau      | 0.1, 1, 5, 10, 20, 50, 100 |
| epsilon  | 10 m |
| noccasions | 100 |
| buffer   | 500 m |
| nx       | 32 |
| nrepl    | 200 |

Here 'sigma' is the overall scale of the O-U distribution and 'tau' determines the serial correlation. An animal is considered to be detected when it is within 'epsilon' of a detector at one of 'noccasions' sampling occasions.

## Package version, date and platform

```{r startup, warning = FALSE, message = FALSE, results = "hold"}
source('../setup.R')
runsim  <- FALSE
figures <- TRUE
cache   <- TRUE
nrepl   <- 200
metadata(runsim) # if FALSE picks up saved metadata.RDS
```

## Changes

2025-07-10 recoded to use simOU functions in secr 5.2.5  
2025-07-10 'grdevice' is set in setup.R to 'ragg_png' allowing graphics on NeSI  

## Simulation code

From secr 5.2.5 it is possible to simulate OU movement and detection directly in secr. Detection happens when the location of an animal at one time step is within epsilon of a detector. The output capthist is a single-occasion sample.

```{r OU initial, eval = TRUE, message = FALSE, results = "hide"}
grid   <- make.grid(nx = 8, ny = 8, spacing = 100, detector = "count")
msk    <- make.mask(grid, buffer = 500, type = 'trapbuffer')
nstep  <- 100
fitarg <- list(detectfn = 'HHN', binomN = 1)
taulevels <- c(0.1, 1, 5, 10, 20, 50, 100)
scen <- make.scenarios( D = 2, epsilon = 10, sigma = 100, tau = taulevels,
                        noccasions = nstep, detectfn = 20)
```

```{r OUcode, eval = runsim, cache = cache, warning = FALSE, message = FALSE}
OUsims <- run.scenarios(
    nrepl       = nrepl, 
    scenarios   = scen, 
    trapset     = grid, 
    maskset     = msk,
    fit         = TRUE, 
    fit.args    = fitarg, 
    seed        = 12345)
saveRDS(OUsims, file = 'OUsims.RDS')
```

```{r OU readRDS, echo = FALSE}
OUsims <- readRDS(file = 'OUsims.RDS')
```

OUsims completed with `r nrepl` replicates in `r round(OUsims$proctime/60,1)` minutes.

## Results

First display some examples of movement tracks.
```{r OU demo1, eval = figures, dev = grdevice, results = "hide", fig.width=7, fig.height=7, fig.cap = "Figure: Examples of correlated movement paths over 100 time steps. Red dot is activity centre; grey circle is 95% contour of activity."}
par(mfrow=c(2,2), mar = c(1,1,2,1), pty='s')
set.seed(123)
for (tau in c(0.1,1,10,100)) {
    locs <- simOU(matrix(c(0,0), nrow=1), tau = tau, sigma = 100, noccasions = 100)
    plot (locs, type='o', xlim=c(-350,350), ylim = c(-350,350), axes=F, xlab='', ylab='')
    mtext(side = 3, paste('tau = ', tau))
    points(0,0, pch = 16, col='red', cex = 1.2)
    box()
    symbols(0,0,circles=245,inches=FALSE, add = TRUE, fg = grey(0.8))
}
```

```{r OU demo2, eval = FALSE, dev = grdevice, echo = FALSE}
# unused code for a whole population of autocorrelated walks
popn <- sim.popn(D = 1, core = grid, buffer = 500)
locs <- simOU(popn, tau = 100, sigma = 100, noccasions = 100)
par(mfrow=c(1,1), mar=c(1,1,1,1))
plot(popn)
mapply(points, locs, type = 'o', col = 1:length(locs))
plot(grid, add=T)
```

```{r OU rpsv, eval = figures, dev = grdevice, cache = TRUE, echo = FALSE, fig.width=6, fig.height=6, fig.cap = "Figure: Bivariate normal sigma for varying tau and n.steps"}
RPSVxy <- function(xy) {
    n <- nrow(xy)
    ss <- function (x) sum(x^2) - (sum(x))^2/n
    sqrt(sum(apply(xy,2,ss))/(2 * (n-2)))
}
set.seed(12345)
popn <- sim.popn(D = 1, core = grid, buffer = 500)  # a set of AC
onetau <- function (tau, n.steps) {
    locs <- simOU(popn, tau = tau, sigma = 100, noccasions = n.steps)
    mean(sapply(locs, RPSVxy))
}
tauval <- 10^seq(-0.5,4,0.5)
steps <- c(20, 100, 500)
par(mfrow = c(1,1), mar = c(5,4,4,4))
plot(1,1, type = 'n', xlim = range(tauval), log = 'x', ylim = c(0,120), xlab = 'tau', ylab = 'Sigma-hat')
for (i in 1:length(steps)) {
    out <- sapply(tauval, onetau, steps[i])
    lines(tauval, out, pch = 15+i, type = 'b')
}
legend(800, 110, legend = steps, lty = 1, pch = 16:18, title = 'n.steps')
```

Check the mean sample sizes for each scenario
```{r OUdatasummary, eval = TRUE, echo = FALSE, results = 'hold'}
counts <- t(sapply(OUsims$output, function(x) apply(sapply(x, attr, 'counts'),1,mean)))
cbind(tau = taulevels, counts[,1:4])
df <- data.frame(Variable = c('tau','n','r','nmov','dpa'), 
           Definition = c('Autocorrelation parameter', 
                          'Number of inndividuals', 
                          'Number of recaptures', 
                          'Number of movements', 
                          'Detectors per individual'))
print(df, row.names = FALSE, right = FALSE)
```

```{r OU tables, eval = TRUE, echo = FALSE}
options(width=100)
cat('Density\n')
cbind(tau = taulevels, round(estimateSummary(OUsims)[,-1],4))
cat('lambda0\n')
cbind(tau = taulevels, round(estimateSummary(OUsims, 'lambda0', 
    statistics = c("true", "nvalid", "EST", "seEST"))[,-1],4))
cat('sigma\n')
cbind(tau = taulevels, round(estimateSummary(OUsims, 'sigma')[,-1],4))
```

```{r OUfigures, eval = figures, dev = grdevice, echo = FALSE, fig.height = 3, fig.width=8, out.width="90%", fig.cap = "Figure: Relative bias RB, relative standard error RSE, and coverage of 95% confidence intervals COV for null-model density estimates from Ornstein-Uhlenbeck model with autocorrelation parameter tau."}
OUsims <- readRDS(file = 'OUsims.RDS')
estD <- estimateSummary(OUsims)[,-1]
estS <- estimateSummary(OUsims, 'sigma')[,-1]
taulevels <- c(0.1, 1, 5, 10, 20, 50, 100)[-1]
par(mfrow=c(1,3), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o', pty='s')
plot(0.1,0,type='n', xlim=c(1,100), log='x', ylim=c(-0.6,0.4), xlab = expression(tau), ylab='RB')
abline(h=0, lty=2)
addRB(taulevels, estD[-1,], type='o', pch=16)
addRB(taulevels, estS[-1,], type='o', pch=1)
legend(1.2, -0.39, legend=c('D','sigma'), pch=c(16,1))

plot(0.1,0,type='n', xlim=c(1,100), log='x', ylim=c(0,0.2), xlab = expression(tau), ylab='RSE')
points(taulevels, estD$RSE[-1], type='o', pch=1)

plot(0.1,0,type='n', xlim=c(1,100), log='x', ylim=c(0.5,1), xlab = expression(tau), ylab='COV')
points(taulevels, estD$COV[-1], type='o', pch=1)
abline(h=0.95, lty = 2)
```

## Interpretation

In these simulations autocorrelation was damaging ($|\mbox{RB}(\hat D)| > 0.05$ only when $\tau \ge 50$. This suggests that autocorrelation *per se* is not a problem; it is possible some other effect comes into play for large $\tau$.

## One-off

Run this code to obtain results for a single scenario. 
```{r OU oneoff, eval = FALSE, message = FALSE, warning = FALSE, results = "hide"}
source('../setup.R')
grid   <- make.grid(nx = 8, ny = 8, spacing = 100, detector = "count")
msk    <- make.mask(grid, buffer = 500, type = 'trapbuffer')
fitarg <- list(binomN = 1, detectfn = 'HHN')
scen   <- make.scenarios( D = 2, epsilon = 10, sigma = 100, tau = 100, 
                        noccasions = 100, detectfn = 20)
oneoff <- run.scenarios(nrepl = 10, scen, trapset = grid,  maskset = msk,
    fit = TRUE, fit.args = fitarg, seed = 12345)
```

