---
title: "**secr** closure simulations"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Closure CLO


The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

## Problem description

Effect of non-closure due to *in situ* births and deaths on density estimated by SECR for studies of different duration. Activity centres fixed.

## Settings

Square grid of 64 binary proximity detectors at 3000-m spacing.  

The same conditions were used for simulation and fitting. 
Weekly 'sessions' were joined in `run.scenarios` for the final single-session analysis.

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| D | 0.001 (0.1/km^2^) |
| lambda0  | 0.2 per week |
| sigma    | 2000 m |
| lambda   | 1.0 (no trend in density) |
| phi      | 0.5 per annum |
| noccasions | 4-24 weeks |
| buffer   | 10000 m |

## Package version, date and platform

```{r startup, warning = FALSE, message = FALSE, results = "hold"}
source('../setup.R')
runsim  <- FALSE
figures <- TRUE
cache   <- TRUE
nrepl   <- 1000
metadata(runsim) # if FALSE picks up saved metadata.RDS
```

## Simulation code

```{r CLO initial, eval = TRUE, message = FALSE, results = "hide"}
nsess  <- seq(4,24,4)
phi    <- c(0.5,0.75, 1.0)
phiw   <- exp(log(phi)/52)   # convert annual to weekly survival
grid   <- make.grid(nx = 8, ny = 8, spacing = 3000, detector = "proximity")
poparg <- expand.arg(
    buffer     = 10000, 
    D          = 0.001, 
    nsessions  = nsess, 
    lambda     = 1, 
    phi        = phiw, 
    sublist = list(details=c('lambda','phi')))

# use lambda0 = 0.2 as approx weekly equiv daily 0.03
detarg <- list(traps = grid, detectfn = 'HHN', detectpar = list(lambda0 = 0.2, sigma = 2000),
               renumber = FALSE, noccasions = 1)
fitarg <- list(mask = make.mask(grid, buffer = 10000, type = 'trapbuffer'), detectfn = 'HHN')

scen <- make.scenarios(D = 0.001, noccasions = 1, detectfn = 'HHN', lambda0 = 0.2, 
                        sigma = 2000, popindex=1:length(poparg))
```

```{r CLO code1, eval = runsim, cache = cache, warning = FALSE, message = FALSE}
CLOsims <- run.scenarios(
    nrepl     = nrepl, 
    scenarios = scen, 
    trapset   = list(grid), 
    pop.args  = poparg,
    det.args  = detarg, 
    joinsessions = TRUE, 
    fit       = TRUE, 
    fit.args  = fitarg, 
    seed      = 12345)
saveRDS(CLOsims, file = 'CLOsims.RDS')
```

## Results

```{r CLO results1, echo = FALSE} 
CLOsims <- readRDS('CLOsims.RDS')
```

CLOsims completed with `r nrepl` replicates in `r round(CLOsims$proctime/60,1)` minutes.

```{r CLO results2, echo = FALSE} 
outraw  <- getcounts(CLOsims)
out  <- summary(CLOsims)$OUTPUT
Dhat <- sapply(out, '[', 'estimate','mean')
RB   <- sapply(out, '[', 'RB','mean')
RSE  <- sapply(out, '[', 'RSE','mean')
COV  <- sapply(out, '[', 'COV','mean')
n <- outraw[,'n']
r <- outraw[,'r']
m <- outraw[,'nmov']
```

```{r CLO tables1, echo = FALSE}
data.frame(Days = nsess*7, Survival = phiw^52, n=n, r=r, m=m, RB=RB, RSE=RSE, COV = COV)
```

Fitting a linear regression to RB as a function of duration we get:
```{r CLO coef, echo = FALSE}
phi0 <- 1:6
phi1 <- 7:12
phi2 <- 13:18
model0 <- lm(I(RB[phi0]) ~ I(nsess*7))
model1 <- lm(I(RB[phi1]) ~ I(nsess*7))
model2 <- lm(I(RB[phi2]) ~ I(nsess*7))
df <- data.frame(phi, rbind(coef(model0), coef(model1), coef(model2)))
names(df) <- c('survival', 'intercept', 'slope')
df$dailymort <- 1 - phi^(1/365)
df$ratio <- df$slope / df$daily
```

```{r CLO figures1, eval = figures, echo = FALSE, fig.height = 3, fig.width=7, out.width="100%", fig.cap = "Relative bias RB, relative standard error RSE, and coverage of 95% confidence intervals COV for density estimates from studies of varying duration. Annual survival 50% (open circles), 75% (filled circles) or 100% (triangles). Horizontal lines for reference only."}
par(mfrow=c(1,3), mar=c(3,4,1,1), pty='s')
plot(0,0,type='n', ylim=c(-0.05,0.3), xlim=c(0,365/2), xlab = 'Duration (days)', ylab='RB', xaxs = 'i')
abline(model0, col='grey')
abline(model1, col='grey')
abline(model2)
points(nsess*7, RB[phi0])
points(nsess*7, RB[phi1], pch = 16)
points(nsess*7, RB[phi2], pch = 2)
abline(h=0, lty=2, col='grey')

plot(0,0,type='n',ylim=c(0,0.3), xlim=c(0,365/2), xlab = 'Duration (days)', ylab='RSE', xaxs = 'i')
points (nsess*7, RSE[phi0])
points (nsess*7, RSE[phi1], pch=16)
points(nsess*7, RSE[phi2], pch = 2)
abline(h=0.1, col='grey')

plot(0,0,type='n',ylim=c(0,1), xlim=c(0,365/2), xlab = 'Duration (days)', ylab='COV', xaxs = 'i')
points(nsess*7, COV[phi0])
points(nsess*7, COV[phi1], pch = 16)
points(nsess*7, COV[phi2], pch = 2)
abline(h=0.95, lty=2, col='grey')
```

## Interpretation

Turnover in a constant population results in positive bias in density estimates on the order of the mortality rate. The daily increase in bias is approximately 70\% of the daily mortality ($1 - \phi^{1/365}$).

Increasing sample size resulted in a small reduction in the estimated sampling variance, but improvement was negligible after 12 weeks. Conversely, the coverage of confidence intervals was OK for durations up to 12 weeks ($\phi = 0.75$) or 8 weeks ($\phi = 0.5$), but poor thereafter.

```{r CLO df, echo = FALSE}
round(df,5)
```

Results are specific to a population with no upward or downward trend.

## One-off

Run this code to obtain results for a single scenario. Requires secrdesign $\ge$ 2.9.0.
```{r CLO oneoff, eval = FALSE, message = FALSE, warning = FALSE, results = "hide"}
source('../setup.R')
nsess  <- 12  # weeks
phi    <- 0.5  # annual survival
phiw   <- exp(log(phi)/52)   # convert annual to weekly survival
grid   <- make.grid(nx = 8, ny = 8, spacing = 3000, detector = "proximity")
poparg <- list(core = grid, buffer = 10000, D = 0.001, nsessions = nsess, 
               details = list(lambda = 1, phi = phiw))
detarg <- list(traps = grid, detectfn = 'HHN', detectpar = list(lambda0 = 0.2, sigma = 2000),
               renumber = FALSE, noccasions = 1)
fitarg <- list(mask = make.mask(grid, buffer = 10000, type = 'trapbuffer'), detectfn = 'HHN')
scen   <- make.scenarios( D = 0.001, noccasions = 1, detectfn = 'HHN', lambda0 = 0.2, sigma = 2000)
sims1  <- run.scenarios(nrepl = 100, scen, trapset = list(grid), pop.args = poparg, det.args = detarg, 
                      joinsessions = TRUE, fit = TRUE, fit.args = fitarg, seed = 12345)
summary(sims1)
```
