---
title: "**secr** simulations ARR"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Array dimensions ARR

The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

A detector array that is too small will likely result in biased estimates. Simulations were provided for searched areas by Efford (2011 Table 3).

These simulations extend the empirical basis for the conclusion.


## Settings

Square grid of 64 binary proximity detectors at spacing $c = 2$.  

The same conditions were used for simulation and fitting. 

| Parameter | Value(s) |
|:-----|:-----------------------|
| detectfn | 'HHN' |
| D | $12.5 (c/2)^{-2}$ |
| lambda0  | 0.1 |
| sigma    | variable |
| noccasions | 10 |
| buffer   | $4\sigma$ |
| nx       | 64 |

## Package version, date and platform

```{r versions, results = 'hold', echo = FALSE}
source('../setup.R')
runsim <- TRUE
metadata(runsim) # if FALSE picks up saved metadata.RDS
```

## Simulation code

### Define ARR scenarios

```{r ARR scenarios, eval = TRUE, message = FALSE, results = "hide"}
grid <- make.grid(nx = 8, ny = 8, spacing = 40, detector = "proximity")
fitarg <- expand.arg(detectfn = c(14,16))
ratio <- seq(0.5,2.5,0.25)
R95 <- max(dist(grid)) / ratio
scen <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, 
    lambda0 = 0.1, sigma = R95/2/2.45, fitindex = 1:2)
# scale density to avoid variation in # detections due to sigma
# result corresponds to D = 12.5 when sigma = 20 (outside simulated range)
scen$D <- scen$D / (scen$sigma / 20)^2
```

```{r ARR scenario summary, cache = TRUE}
scenarioSummary(scen, grid)
```

### Run ARR simulations

```{r ARR run, eval = runsim, cache = TRUE}
ARRsims <- run.scenarios(nrepl = 500, scen, trapset = grid, nx = 64,
    fit = TRUE, fit.args = fitarg, seed = 12345)
saveRDS(ARRsims, file = 'ARRsims.RDS')
```

## Results

```{r ARR results, eval = TRUE, echo = TRUE}
ARRsims <- readRDS(file = 'ARRsims.RDS')
estD <- estimateSummary(ARRsims)
estD
```

Note: rRMSE is the root-mean-square error divided by the true density.

$\mbox{rRMSE}(\hat D) = \frac{1}{D} \sqrt {\frac{1}{N} \sum_{i=1}^N {(D - \hat D_i)^2}}$ 

$\mbox{RB}(\hat D) = \frac{1}{N} \sum_{i=1}^N {(D - \hat D_i)}/D$ 


```{r ARR figure, eval = TRUE, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Effect of array size on relative bias (RB) and relative root-mean-square error (rRMSE) of density estimates"}

par(mfrow=c(1,2), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o', cex = 1, pty='s')

x <- max(dist(grid)) / (2.45 * scen$sigma[1:9] *2)

plot(0,0,type='n', xlim=c(0,2.8), ylim=c(-0.2, 0.2), 
     xlab = 'Array diameter / HR diameter', 
     ylab = expression(paste('RB (', hat(italic(D)), ')') ) )
shade(0.05)
abline(h=0, lty=2)
abline(v=0, col='grey')
addRB(x, estD[1:9,], type='o', pch=16, cex=1.2)
addRB(x, estD[10:18,], type='o', pch=24, bg = 'white', xoffset = 0.02)
legend(2, 0.2, legend=c('HHN','HEX'), pch=c(16,24), pt.bg = 'white', cex = 1)

plot(0,0,type='n', xlim=c(0,2.8), ylim=c(0, 0.5), 
     xlab = 'Array diameter / HR diameter',
     ylab = expression(paste('rRMSE (', hat(italic(D)), ')') ) )
points(x, estD[1:9,'rRMSE'], type='o', pch=16, cex = 1.2)
points(x+0.02, estD[10:18,'rRMSE'], type='o', pch=24, bg='white', cex=1.1)
legend(2, 0.5, legend=c('HHN','HEX'), pch=c(16,24), pt.bg = 'white', cex = 1)
```

## Compare Efford 2011 Table 3

Efford (2011 Ecology 92: 2202--2207) simulated area-search data with a half-normal detection function. The searched area varied from $1 \sigma^2$ to $64 \sigma^2$. We plot the results as a function of area-diameter divided by 95\% BVN home range as with the preceding simulations. Bias results when the linear diagonal of the searched area is less than about 75\% of the 95\% home-range diameter. Given the arbitrariness of both measures (area dimension, home range size) we suggest the simple rule: the grid or searched area should be at least the size of the home range.

```{r plotefford2011}
searcharea <- c(1,2,4,8,16,32,64)
diag <- sqrt(2) * sqrt(searcharea)
HR95 <- 2 * 2.45 * 1
diag/HR95
RB <- c(0.47,0.20,0.12,0.02, 0.02, 0.02,-0.01)

par(mfrow=c(1,1), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o', cex = 1.2, pty='s')

plot(0,0,type='n', xlim=c(0,2.8), ylim=c(-0.2, 0.5), 
     xlab = 'Array diameter / HR diameter', 
     ylab = expression(paste('RB (', hat(italic(D)), ')') ) )
shade(0.05)
abline(h=0, lty=2)
addRB(x, estD[1:9,], type='o', pch=16, cex=1)
points(diag/HR95, RB, pch=21, bg='yellow')
```

