---
title: "**secr** simulations MID"
author: "Murray Efford"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage[utf8]{inputenc}
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: 4
  word_document: default
---

# Mis-identification as source of bias in SECR

The simulations described here are part of a series documenting the effect of 
breaching assumptions of spatially explicit capture--recapture models.

## Problem description

Detected animals may be 

1. mis-identified as an existing marked (known) individual, or 
2. mistakenly identified as a new 'ghost' individual, or
3. indistinguishable from another with the same genotype at the loci sampled, or
4. mistakenly tagged as a new individual after loss of a previous tag.

We simulate these mechanisms separately (MID1, MID2, MID3, MID6).

## Settings

MID1-5 Square grid of 64 binary proximity detectors at spacing $c = 2\sigma$.  
MID6 Square grid of 64 multi-catch traps at spacing $c = 2\sigma$.  

The same conditions were used for simulation and fitting.

| Parameter || Value(s) |
|:-----|-|:----------------------|
| detectfn || 'HHN' |
| D        || $12.5 \sigma^{-2}$ |
| lambda0  || 0.1 |
| sigma    || c/2 |
| noccasions || 10 |
| buffer   || $4\sigma$ |
| nx       || 32 |

## Package version, date and platform

```{r startup, warning = FALSE, message = FALSE, results = "hold"}
source('../setup.R')
runsim  <- FALSE
figures <- TRUE
cache   <- TRUE
nrepl   <- 200
metadata(runsim) # if FALSE picks up saved metadata.RDS
```

## Simulation code

This alternative capthist simulation function generates a capthist the usual 
way with `sim.capthist` and then scrambles the identities according to one of 
the four mechanisms.

```{r MID capthist, eval = TRUE, message = FALSE, results = "hide"}
simMID.capthist <- function(traps, popn, detectfn, detectpar = list(),
                            noccasions = 10, pMID = 0, pGhost = 0, PI = 0, 
                            pTagloss = 0, ...)
{
    ch0 <- sim.capthist(traps, popn, detectfn, detectpar, noccasions)
    n <- nrow(ch0)
    capt <- as.data.frame(ch0)
    capt$ID <- as.integer(capt$ID)
    newcapt <- capt
    if (pMID > 0) {
        # on any occasion, ID reassigned to any known animal with probability pMID
        whichmisID <- runif(nrow(capt)) < pMID
        newID <- sample.int(n, size = sum(whichmisID), replace = TRUE)
        newcapt$ID[whichmisID] <- newID
    }
    else if (pGhost>0) {
        # proportion pGhost of all detections receive unrepeatable ID
        novel <- runif(nrow(capt)) < pGhost
        newcapt$ID[novel] <- paste('N',capt$ID[novel], capt$Occasion[novel], capt$TrapID[novel], sep='.')
    }
    else if (PI>0) {
        pairwise <- 1 - matrix(runif(n*n) < PI, nrow = n)
        # find any pairs, triplets etc. with zero distance and assign common ID
        d <- as.dist(pairwise)   # uses only lower triangle of pairwise
        groupID <- cutree(hclust(d, 'single'), h = 0.1)  # groups individuals linked by dist==0
        newcapt$ID <- groupID[capt$ID]
    }
    else if (pTagloss>0) {
        # detections receive new ID with probability dependent on interval from last
        elapsed <- c(NA, diff(capt$Occasion))           # assume sorted by ID, occasion
        elapsed[match(unique(capt$ID), capt$ID)] <- NA  # not first
        novel <- runif(nrow(capt)) < (1 - (1-pTagloss)^elapsed)
        novel[is.na(novel)] <- FALSE
        novelist <- split(novel, capt$ID)  # assumes ID ordered factor levels
        subID <- unlist(lapply(novelist, cumsum))
        newcapt$ID <- paste('N',capt$ID, subID, sep='.')
    }
    # re-read
    ch <- make.capthist(newcapt, traps, noccasions = noccasions)
    if (PI>0) {
        identicalpairs <- cbind(as.numeric(row(pairwise)), as.numeric(col(pairwise)))[!as.logical(pairwise),]
        attr(ch, 'identicalpairs') <- identicalpairs
        attr(ch, 'groupID') <- groupID
    }
    ch
}
```

```{r common settings, eval = TRUE, message = FALSE, results = "hide"}
grid <- make.grid(nx = 8, ny = 8, spacing = 40, detector = "proximity")
fitarg <- list(detectfn = 'HHN', model = lambda0 ~ 1, start = NULL)
```

### Run MID simulations

```{r test simMID.capthist, eval = FALSE, echo = FALSE, message = FALSE, results = "hide"}
# test code; large lambda0
pop <- sim.popn(12.5, grid)
ch <- simMID.capthist (grid, popn = pop, detectfn=14, detectpar = list(lambda0=0.2, sigma=20),
                       noccasions = 10, pMID=0, pGhost=0.1)
summary(ch)
ch <- simMID.capthist (grid, popn = pop, detectfn=14, detectpar = list(lambda0=0.2, sigma=20),
                       noccasions = 10, PI=0.001)
summary(ch)
```

```{r MID1-scen, eval = TRUE, message = FALSE, warning = FALSE, results = "hide"}
detarg1 <- expand.arg(pMID = c(0.00, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2))
scen1 <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, 
    lambda0 = 0.1, sigma = 20, detindex = 1:length(detarg1), fitindex = 1)
```

```{r MID1 scenario summary, cache = cache}
scenarioSummary(scen1, grid)
```

```{r MID1-run, eval = runsim, cache = cache, message = FALSE, warning = FALSE, results = "hide"}
MIDsims1 <- run.scenarios(
    nrepl     = nrepl, 
    scenarios = scen1, 
    trapset   = grid, 
    det.args  = detarg1, 
    fit       = TRUE, 
    fit.args  = fitarg, 
    seed      = 1234, 
    CH.function = "simMID.capthist")
saveRDS(MIDsims1, file = 'MIDsims1.RDS')
```

```{r MID2-scen, eval = TRUE, message = FALSE, warning = FALSE, results = "hide"}
detarg2 <- expand.arg(pGhost = c(0.00, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2))
scen2 <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, 
    lambda0 = 0.1, sigma = 20, detindex = 1:length(detarg2), fitindex = 1)
```

```{r MID2 scenario summary, cache = cache}
scenarioSummary(scen2, grid)
```

```{r MID2-run, eval = runsim, cache = cache, message = FALSE, warning = FALSE, results = "hide"}
MIDsims2 <- run.scenarios(
    nrepl       = nrepl, 
    scenarios   = scen2, 
    trapset     = grid, 
    det.args    = detarg2, 
    fit         = TRUE, 
    fit.args    = fitarg, 
    seed        = 1234, 
    CH.function = "simMID.capthist")
saveRDS(MIDsims2, file = 'MIDsims2.RDS')
```

```{r MID3-scen, eval = TRUE, message = FALSE, warning = FALSE, results = "hide"}
detarg3 <- expand.arg(PI = c(0.00, 10^(seq(-5,-2,0.5))))
scen3 <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, 
    lambda0 = 0.1, sigma = 20, detindex = 1:length(detarg3))
```

```{r MID3 scenario summary, cache = cache}
scenarioSummary(scen3, grid)
```

```{r MID3-run, eval = runsim, cache = cache, message = FALSE, warning = FALSE, results = "hide"}
MIDsims3 <- run.scenarios(
    nrepl       = nrepl, 
    scenarios   = scen3, 
    trapset     = grid, 
    det.args    = detarg3, 
    fit         = TRUE, 
    fit.args    = fitarg, 
    seed        = 1234, 
    CH.function = "simMID.capthist")
saveRDS(MIDsims3, file = 'MIDsims3.RDS')
```

```{r MID4-scen, eval = TRUE, message = FALSE, warning = FALSE, results = "hide"}
scen4 <- make.scenarios(D = 12.5 * c(0.25,0.5,1,2,4), noccasions = 10, detectfn = 14, 
    lambda0 = 0.1, sigma = 20, detindex = 1:length(detarg3))
scen4 <- scen4[order(scen4$D, scen4$detindex),]
scen4$scenario <- 1:nrow(scen4)
```

```{r MID4 scenario summary, cache = cache}
scenarioSummary(scen4, grid)
```

```{r MID4-run, eval = runsim, cache = cache, message = FALSE, warning = FALSE, results = "hide"}
MIDsims4 <- run.scenarios(
    nrepl       = nrepl, 
    scenarios   = scen4, 
    trapset     = grid, 
    det.args    = detarg3, 
    fit         = TRUE,  
    fit.args    = fitarg, 
    seed        = 1234, 
    CH.function = "simMID.capthist")
saveRDS(MIDsims4, file = 'MIDsims4.RDS')
```

```{r MID5-scen, eval = TRUE, message = FALSE, warning = FALSE, results = "hide"}
scen5 <- make.scenarios(D = 12.5, noccasions = 10, detectfn = 14, 
    lambda0 = c(0.05,0.1,0.2), sigma = 20, detindex = 1:length(detarg3))
scen5 <- scen5[order(scen5$lambda0, scen5$detindex),]
scen5$scenario <- 1:nrow(scen5)
```

```{r MID5 scenario summary, cache = cache}
scenarioSummary(scen5, grid)
```

```{r MID5-run, eval = runsim, cache = cache, message = FALSE, warning = FALSE, results = "hide"}
MIDsims5 <- run.scenarios(
    nrepl       = nrepl, 
    scenarios   = scen5, 
    trapset     = grid, 
    det.args    = detarg3, 
    fit         = TRUE, 
    fit.args    = fitarg, 
    seed        = 1234, 
    CH.function = "simMID.capthist")
saveRDS(MIDsims5, file = 'MIDsims5.RDS')
```

```{r MID6-scen, eval = TRUE, message = FALSE, warning = FALSE, results = "hide"}
detarg6 <- expand.arg(pTagloss = c(0.00, 0.002, 0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2))
scen6 <- make.scenarios(D = 12.5, noccasions = c(5,10), detectfn = 14, 
    lambda0 = 0.1, sigma = 20, detindex = 1:length(detarg6), fitindex = 1)
gridmulti <- make.grid(nx = 8, ny = 8, spacing = 40, detector = "multi")
```

```{r MID6 scenario summary, cache = cache}
scenarioSummary(scen6, gridmulti)
```

```{r MID6-run, eval = runsim, cache = cache, message = FALSE, warning = FALSE, results = "hide"}
MIDsims6 <- run.scenarios(
    nrepl     = nrepl, 
    scenarios = scen6, 
    trapset   = gridmulti, 
    det.args  = detarg6, 
    fit       = TRUE, 
    fit.args  = fitarg, 
    seed      = 1234, 
    CH.function = "simMID.capthist")
saveRDS(MIDsims6, file = 'MIDsims6.RDS')
```


## Results

The simulation scenario resulted in the detection of about 110 individuals in the null case (scenario 1 with pMID = 0 etc.).

### MID1 Random mis-identification

```{r MID-tables1, eval = TRUE, echo = FALSE, results = "hold"}
options(width=110, digits=4)
MIDsims1 <- readRDS(file = 'MIDsims1.RDS')
```

MIDsims1 completed with `r nrepl` replicates in `r round(MIDsims1$proctime/60,1)` minutes.

```{r MID-tables2, eval = TRUE, echo = FALSE}
scenarios <- attr(detarg1, 'comb')
cbind(scenarios, getcounts(MIDsims1)[,-5])
countlegend(-5)
```

```{r MID-tables3, echo = FALSE}
estD <- estimateSummary(MIDsims1)[,-1]
estL <- estimateSummary(MIDsims1, 'lambda0')[,-1]
estS <- estimateSummary(MIDsims1, 'sigma')[,-1]
cat('Density\n')
cbind(scenarios, round(estD,4))
cat('lambda0\n') 
cbind(scenarios, round(estL,4))
cat('sigma\n')
cbind(scenarios, round(estS,4))
```

```{r MID1 figure, eval = figures, echo = FALSE, results = "hide", fig.width = 5, fig.height = 5, fig.cap = "Figure: Relative bias of parameter estimates from null model given misidentification. Shaded band: |RB| < 10%." }
par(mfrow=c(1,1), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o', bg = 'white', pty='s')
x <- scenarios$pMID
# when does RB exceed 10%?
approx(estD$RB, x, xout=-0.1)$y
# [1] 0.023488
leg <- c('D','lambda0','sigma')
plot(0,0,type='n', xlim=c(0.0,0.2), log='', ylim=c(-1,1), xlab = 'Proportion mis-assigned', ylab='RB')
shade(0.1)
abline(h=0, lty=2)
addRB(x, estL, type='o', pch=21, bg = 'white')
addRB(x, estS, type='o', pch=24, bg = 'white', star = 1.0)
addRB(x, estD, type='o', pch=21, bg = yob5[3], cex = 1.3)
legend(0.15, 0.6, legend = leg, pch=c(21,21,24), cex = 0.85, pt.cex=1.2, pt.bg=c(yob5[3],'white','white'))
```

### MID2 Ghost individuals

```{r MID2-table1, eval = TRUE, echo = FALSE}
MIDsims2 <- readRDS(file = 'MIDsims2.RDS')
scenarios <- attr(detarg2, 'comb')
```

MIDsims2 completed with `r nrepl` replicates in `r round(MIDsims2$proctime/60,1)` minutes.

```{r MID2-table2, eval = TRUE}
cbind(scenarios, getcounts(MIDsims2)[,-5])
```

```{r MID2 figure, eval = figures, echo = FALSE, results = "hide", fig.width = 5, fig.height = 5, fig.cap = "Figure: Relative bias of parameter estimates from null model vs frequency of ghost identities. Shaded band: |RB| < 10%" }
estD <- estimateSummary(MIDsims2)[,-1]
estL <- estimateSummary(MIDsims2, 'lambda0')[,-1]
estS <- estimateSummary(MIDsims2, 'sigma')[,-1]
par(mfrow=c(1,1), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o', bg = 'white', pty='s')
x <- scenarios$pGhost
# when does RB exceed 10%?
approx(estD$RB, x, xout=0.1)$y 
# [1] 0.032302
leg <- c('D','lambda0','sigma')
plot(0,0,type='n', xlim=c(0.0,0.2), log='', ylim=c(-1,1), xlab = 'Proportion ghosted', ylab='RB')
shade(0.1)
abline(h=0, lty=2)
addRB(x, estL, type='o', pch=21, bg = 'white')
addRB(x, estS, type='o', pch=24, bg = 'white', star = 1.0)
addRB(x, estD, type='o', pch=21, bg = yob5[3],cex = 1.3)
legend(0, 0.-0.5, legend = leg, pch=c(21,21,24), cex = 0.85, pt.cex=1.2, pt.bg=c(yob5[3],'white','white'))
```

### MID3 Identical genotypes

```{r MID3-table1, eval = TRUE, echo = FALSE}
MIDsims3 <- readRDS(file = 'MIDsims3.RDS')
scenarios <- attr(detarg3, 'comb')
```

MIDsims3 completed with `r nrepl` replicates in  `r round(MIDsims3$proctime/60,1)` minutes.

```{r MID3-table2, eval = TRUE, echo = FALSE}
cbind(scenarios, getcounts(MIDsims3)[,-5])
estD <- estimateSummary(MIDsims3)
estL <- estimateSummary(MIDsims3, 'lambda0')
estS <- estimateSummary(MIDsims3, 'sigma')
cat('Density\n')
cbind(scenarios, round(estD,4))
cat('lambda0\n') 
cbind(scenarios, round(estL,4))
cat('sigma\n')
cbind(scenarios, round(estS,4))
```

```{r MID3 figure, eval = figures, echo = FALSE, results = "hide", fig.width = 6, fig.height = 6, fig.cap = "Figure: Relative bias of parameter estimates from null model when some individuals cannot be distinguished. Shaded band: |RB| < 10%. Inset shows same plot with arithmetic x-axis." }
par(mfrow=c(1,1), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o', bg = 'white', pty='s', fig = c(0,1,0,1))
x <- scenarios$PI
# when does RB exceed 10%?
approx(estD$RB, x, xout=-0.1)$y 
# [1] 0.00037487
leg <- c('D','lambda0','sigma')
plot(1,1,type='n', xlim=c(1e-5, 1e-2), log='x', ylim=c(-1,1), xlab = 'Probability of identity PI (log scale)', ylab='RB', axes = F)
axis(1, at = 10^seq(-5,-2,0.5), label = c('0.00001','','0.0001','','0.001','','0.01'))
axis(2)
shade(0.1)
abline(h=0, lty=2)
addRB(x, estL, type='o', pch=21, bg = 'white')
addRB(x, estS, type='o', pch=24, bg = 'white', star = 1.0)
addRB(x, estD, type='o', pch=21, bg = yob5[3],cex = 1.3)
legend(0.000015, -0.6, legend = leg, pch=c(21,21,24), cex = 0.85, pt.cex=1.2, pt.bg=c(yob5[3],'white','white'))

par(fig = c(0.09,0.4, 0.6, 0.98), mgp=c(0.5,0.5,0), new = T)  
plot(1,1,type='n', xlim=c(0, 1e-2),  ylim=c(-1,1), xlab = 'PI', ylab='RB', axes = F)
shade()
abline(h=0, lty=2)
addRB(x, estL, type='o', pch=21, bg = 'white')
addRB(x, estS, type='o', pch=24, bg = 'white')
addRB(x, estD, type='o', pch=16)
axis(1, at = 10^seq(-5,-2,0.5), label = FALSE)
```

### MID4 Identical genotypes, varying density

```{r MID4-table1, eval = TRUE, echo = FALSE}
MIDsims4 <- readRDS(file = 'MIDsims4.RDS')
scenarios <- expand.grid(PI = c(0.00, 10^(seq(-5,-2,0.5))), D = 12.5 * c(0.25,0.5,1,2,4))
```

MIDsims4 completed with `r nrepl` replicates in `r round(MIDsims4$proctime/60,1)` minutes.

```{r MID4-table2, eval = TRUE, echo = FALSE}
cbind(scenarios, getcounts(MIDsims4)[,-5])
countlegend(-5)
estD <- estimateSummary(MIDsims4)
estL <- estimateSummary(MIDsims4, 'lambda0')
estS <- estimateSummary(MIDsims4, 'sigma')
cat('Density\n')
cbind(scenarios, round(estD,4))
cat('lambda0\n') 
cbind(scenarios, round(estL,4))
cat('sigma\n')
cbind(scenarios, round(estS,4))
```

```{r MID4 figure, eval = figures, echo = FALSE, results = "hide", fig.width = 6, fig.height = 6, fig.cap = "Figure: Relative bias of parameter estimates from null model when some individuals cannot be distinguished. Shaded band: |RB| < 10%. " }
par(mar = c(4,4,1,1), mgp = c(2.4,0.7,0), bty = 'o', bg = 'white', pty = 's')
x <- scenarios$PI[1:8]
plot(1,1,type='n', xlim=c(1e-5, 1e-2), log='x', ylim=c(-1,0.5), 
     xlab = 'Probability of identity PI (log scale)', ylab='RB', axes = F)
axis(1, at = 10^seq(-5,-2,0.5), label = c('0.00001','','0.0001','','0.001','','0.01'))
axis(2)
shade(0.1)
abline(h = 0, lty = 2)
for (d in 1:5) {
    addRB(x, estD[(d-1)*8 + 1:8,], type = 'o', pch = 21, cex = 1.4, bg = yob5[d])
}
legend(0.000015, -0.55, legend = as.character(12.5 * c(0.25,0.5,1,2,4)),
       text.width = 0.38, title = expression(paste('Density ', sigma^{-2})), 
       pch = 21, pt.bg = yob5, cex=0.85, pt.cex = 1.4)
```

From the curves in the previous figure we can interpolate the threshold PI at which relative bias of the density estimate from a null model exceeds a given threshold e.g. 5\%. We plot this as a function of the expected number of individuals detected.

```{r threshold, eval = figures, echo = FALSE}
par(pty = 's')
threshold05 <- sapply(1:5, function(d) approx(estD[(d-1)*8 + 1:8,'RB'], x, xout=-0.05)$y)
threshold10 <- sapply(1:5, function(d) approx(estD[(d-1)*8 + 1:8,'RB'], x, xout=-0.1)$y)
n <- getcounts(MIDsims4)[seq(1,40,8),'n']
plot(1,1,type='n', log='y', xlim=c(0,500), ylim=c(1e-5,1e-2), 
     xlab = expression(paste('Expected number detected ', ~italic(n))), ylab = '', axes = FALSE)
axis(2, at = 10^seq(-5,-2,0.5), label = c('0.00001','','0.0001','','0.001','','0.01'), las = 1)
axis(1)
mtext(side=2, line = 4, 'PI threshold')
box()
points(n, threshold05, pch=21, type = 'o', bg = yob5, cex = 1.5)
points(n, threshold10, pch=21, type = 'o', bg = yob5, cex = 1.5)
text(380, 2.2e-4, 'RB -10%')
text(380, 3e-5, 'RB -5%')
```

### MID5 identical genotypes, varying detection probability

```{r MID5-table1, eval = TRUE, echo = FALSE}
MIDsims5 <- readRDS(file = 'MIDsims5.RDS')
scenarios <- expand.grid(PI = c(0.00, 10^(seq(-5,-2,0.5))), lambda0 = c(0.05,0.1,0.2))
```

MIDsims5 completed with `r nrepl` replicates in `r round(MIDsims5$proctime/60,1)` minutes.

```{r MID5-table2, eval = TRUE, echo = FALSE}
cbind(scenarios, getcounts(MIDsims5)[,-5])
countlegend(-5)
estD <- estimateSummary(MIDsims5)
cat('Density\n')
cbind(scenarios, round(estD,4))
```

```{r MID5 figure, eval = figures, echo = FALSE, results = "hide", fig.width = 6, fig.height = 6, fig.cap = "Figure: Relative bias of parameter estimates from null model when some individuals cannot be distinguished. Shaded band: |RB| < 10%. " }
par(mar = c(4,4,1,1), mgp = c(2.4,0.7,0), bty = 'o', bg = 'white', pty = 's', fig = c(0,1,0,1))
x <- scenarios$PI[1:8]
plot(1,1,type='n', xlim=c(1e-5, 1e-2), log='x', ylim=c(-1,0.5), 
     xlab = 'Probability of identity PI (log scale)', ylab='RB', axes = F)
axis(1, at = 10^seq(-5,-2,0.5), label = c('0.00001','','0.0001','','0.001','','0.01'))
axis(2)
shade(0.1)
abline(h = 0, lty = 2)
for (l in 1:3) {
    addRB(x, estD[(l-1)*8 + 1:8,], type = 'o', pch = 21, cex = 1.4, bg = blu5[l+1])
}
legend(0.000015, -0.55, legend = as.character(c(0.05,0.1,0.2)),
       text.width = 0.38, title = 'lambda0', 
       pch = 21, pt.bg = blu5[2:4], cex=0.85, pt.cex = 1.4)
```

### MID6 Tag loss

```{r MID-tables6a, eval = TRUE, echo = FALSE, results = "hold"}
options(width=110, digits=4)
MIDsims6 <- readRDS(file = 'MIDsims6.RDS')
```

MIDsims6 completed with `r nrepl` replicates in `r round(MIDsims6$proctime/60,1)` minutes.

```{r MID-tables6b, eval = TRUE, echo = FALSE}
scenarios <- cbind(noccasions=scen6$noccasions, attr(detarg6, 'comb')[scen6$detindex,,drop=F])
cbind(scenarios, getcounts(MIDsims6)[,-5])
countlegend(-5)
```

```{r MID-tables6c, echo = FALSE}
estD <- estimateSummary(MIDsims6)[,-1]
estL <- estimateSummary(MIDsims6, 'lambda0')[,-1]
estS <- estimateSummary(MIDsims6, 'sigma')[,-1]
cat('Density\n')
cbind(scenarios, round(estD,4))
cat('lambda0\n') 
cbind(scenarios, round(estL,4))
cat('sigma\n')
cbind(scenarios, round(estS,4))
```

```{r MID6 thresholds}
# when does RB exceed 10%?
OK <- scenarios$noccasions == 5
approx(estD[OK,]$RB, scenarios$pTagloss[OK], xout=0.1)$y
OK <- scenarios$noccasions == 10
approx(estD[OK,]$RB, scenarios$pTagloss[OK], xout=0.1)$y
```

```{r MID6 figure, eval = figures, echo = FALSE, results = "hide", fig.width = 5, fig.height = 5, fig.cap = "Figure: Relative bias of parameter estimates from null model given tag loss. Shaded band: |RB| < 10%." }
OK <- scenarios$noccasions == 10
x <- scenarios$pTagloss[OK]
par(mfrow=c(1,1), mar=c(4,4,1,1), mgp=c(2.4,0.7,0), bty = 'o', bg = 'white', pty='s')
leg <- c('D','lambda0','sigma')
plot(0,0,type='n', xlim=c(0.002,0.2), log='x', ylim=c(-1,1), xlab = 'Proportion tags lost per occasion', ylab='RB')
shade(0.1)
abline(h=0, lty=2)
addRB(x, estL[OK,], type='o', pch=21, bg = 'white')
addRB(x, estS[OK,], type='o', pch=24, bg = 'white', star = 1.0)
addRB(x, estD[OK,], type='o', pch=21, bg = yob5[3], cex = 1.3)
legend(0.003, 0.9, legend = leg, pch=c(21,21,24), cex = 0.85, pt.cex=1.2, pt.bg=c(yob5[3],'white','white'))
```

## Interpretation

### MID1

Random mis-identification of individuals in more than about 2\% of detection events caused significant negative bias in estimates of density from the null model ($RB(\hat D) < -10\%$). The simulation model was only an approximation to study conditions as it *excluded* the possibility that mis-identification placed an individual outside the set of otherwise known individuals (i.e., as a new never-to-be-seen-again ghost individual). 

### MID2

Ghost individuals caused positive relative bias in density estimates on the order of 3 times their frequency of occurrence expressed as a proportion of all detections ($RB(\hat D) > +10\%$ when $\mbox{pGhost}>0.032$). The simulated high frequencies (>5\%) seem rather improbable. Estimates of $\sigma$ remained unbiased (with potentially a slight reduction in precision owing to the reduced number of recaptures).

### MID3

Density estimates were sensitive to the presence of identical genotypes. Any $\mbox{PI} > 0.0004$ resulted in significant negative bias ($RB(\hat D) < -10\%$). Bias is not directly attributable to a reduction in the number of distinct individuals, as this varied little at this threshold. For 110 observed individuals there are 5995 pairwise possible identities, The threshold PI corresponds to an average of just 2.4 pairs

### MID4

Varying true density has a large effect on the relationship between PI and bias in $\hat D$. This effect is probably due to the increased number of individuals that are detected and therefore potentially confused. Parameters of detection ($\lambda_0, \sigma$) also affect the number detected.

### MID5

Although varying detection probability ($\lambda_0$) in general should shift $\mbox E (n)$ and hence $\mbox{RB}(\hat D)$ for given PI, these simulations effected only a small change in $\mbox E (n)$.

### MID6

Tag loss caused positive relative bias in density estimates on the order of the per-occasion probability (pTagloss) times the number of occasions divided by 2. In the simulations with 10 occasions, $RB(\hat D) > +10\%$ when pTagloss $>$ 0.019; the threshold increased to 0.044 when the number of occasions was reduced to 5. Estimates of $\sigma$ remained unbiased (with potentially a slight reduction in precision owing to the reduced number of recaptures).

